### 2025-04-22 paper.tex
Initial commit

\documentclass[
  11pt,
  fleqn
]{article}
% [fleqn] left-aligns all equations

\input{preamble/preamble}

% For writing/planning purposes: show paragraph level in TOC
\setcounter{secnumdepth}{2}

\addbibresource{bibliography.bib}
\graphicspath{{fig/}}

\title{Guidance on choosing ordinal outcomes and endpoints for clinical trials}
\author{Leon Di Stefano, Katherine Lee, etc\ldots}
\date{\today}

\begin{document}

\maketitle

\tableofcontents

\section{Introduction}

Ordinal outcomes have become increasingly prominent in clinical
resarch, particularly following the COVID-19 pandemic.

Some advantages of ordinal outcomes are that they can
\begin{enumerate}
  \item Provide more information that binary outcomes, including
    composite outcomes, allowing for more efficient and ethical studies
  \item Allow for the incorporation of safety events into one's
    primary outcome and therefore account for many different aspects
    of benefit/harm. At the same time, they do not require a
    fully-specified utility function (which may differ from patient
    to patient or require more extensive development and validation)
  \item Allow for the combination of discrete and continous outcome
    measures (for example a numerical quality of life scale with
    death as the worst outcome)
  \item Allow for the incorporation of intercurrent events into one's
    primary outcome, avoiding complexities of alternative estimand strategies
    like principal stratum estimands (particular for truncating
    events like death) while having more statistical efficiency than
    binary composite estimands (see point (1))
\end{enumerate}

The goal of this manuscript is to serve as a guide for clinicians and trialists
in choosing/developing ordinal outcomes for their studies. We consider both
both the outcome itself (the data recorded on each patient) and effect
measures or ``endpoints'' (the metric used to compare patient groups).

We want to distinguish \emph{choosing} an outcome/effect
measure from \emph{validating} the outcome/effect measure using
external data. This paper is about choosing the outcome and effect
measure. In practice choosing an outcome and effect measure and
validating that outcome and effect measure in pre-trial data will be
an iterative process.

\section{Desiderata for outcomes and effect measures}

We want outcomes that
\begin{enumerate}
  \item reflect patient benefit and harm
  \item capture sufficient information to efficiently discriminate
    among better- and
    worse-off patients.
\end{enumerate}

We want effect measures that
\begin{enumerate}
  \item capture decision-relevant differences in patient benefit and harm
  \item are easy to interpret and to translate into practical recommendations
\end{enumerate}

\section{Techniques for constructing ordinal outcomes}

\subsection{Tie-breaking can be used to fine-grain ordinal outcomes}

Two kinds: crossing, and tie breaking within a level (overriding).

\subsection{Category probabilities affect precision and power mostly via the
probability of the largest category}

\subsection{Ordinal outcomes can incorporate longitudinal or
time-to-event information}

Distinguish between ``state'' and ``trajectory'' variables.

Trajectory variables, including times-to-event, summarize over a timecourse.

These can be incorporated into an ordinal outcome. For example ``worst pSOFA
while hospitalized until day 28''.

Arguably most existing time-to-event analyses are in fact analyses of an
ordinal outcome. The Cox model treats time-to-event as an ordinal outcome:
because (1) the baseline hazard is not modelled, (2) the model itself is
invariant to monotone rescalings of the time axis, and (3) the (partial)
likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
marginal likelihood).

\section{Guidelines for choosing estimands or effect measures}

\subsection{Model-based versus nonparametric effect measures}

Model-based: common odds ratio; common stopping ratio or common
continuation ratio.

Nonparametric or model-free measures: probabilistic index, win ratio, win odds,
...

Both of these can be considered ``estimands'', insofar as we consider.

Nonparametric effect measures may appear to be assumption-free. However
\begin{itemize}
  \item Assumptions don't need to hold exactly for the model-based
    effect measures to make sense
  \item The model-free methods also may not make sense, if for
    example there are nonmonotone effects across cutpoints
  \item Model-based methods have additional advantages, for example
    the possibility of adjustment for covariates.

    When additivity approximately holds, this yields a ``best of both
    worlds'' effect measure that is
    personalized effect measure that is yet reportable as a single
    number. Risk ratio for vaccine efficacy (being approximately
    constant as a function of baseline risk/age) as an example.
\end{itemize}

Simulations show a typically very strong association between
proportional odds ratios and probabilistic indices [cite Harrell].

Reducing to a composite binary outcome
\begin{itemize}
  \item Based on a specific cutpoint, e.g. ``hospitalization or worse''
  \item Based on a quantile in a reference group; for example,
    ``worse than the median outcome under standard care''
\end{itemize}

One point of view is that one would ``ideally'' analyze ordinal
outcomes by reporting the associated expected utilities [cite Berry
talk; example of utility-weighted Rankin scale].

\subsection{Generalized pairwise comparisons (``win'' methods)
  are implicitly about constructing
ordinal outcomes, in a way that is compatible with right-censoring}

\end{document}

### 2025-04-23 paper.tex
Lots of text + pmax efficiency figure

@@
 
-\title{Guidance on choosing ordinal outcomes and endpoints for clinical trials}
+\title{Advice on developing ordinal outcomes and endpoints for clinical trials}
 \author{Leon Di Stefano, Katherine Lee, etc\ldots}
@@
 \tableofcontents
+\newpage
 
@@
 
-We want to distinguish \emph{choosing} an outcome/effect
-measure from \emph{validating} the outcome/effect measure using
-external data. This paper is about choosing the outcome and effect
-measure. In practice choosing an outcome and effect measure and
-validating that outcome and effect measure in pre-trial data will be
-an iterative process.
+We want to distinguish \emph{choosing} an outcome/effect measure from
+\emph{validating} the outcome/effect measure using external data. This paper is
+about the first task, choosing (or constructing) the outcome and effect
+measure. In practice choosing an outcome and effect measure and validating that
+outcome and effect measure in pre-trial data will be an iterative process; our
+focus is on the ``positive'' side of the process, namely coming up
+with, and tweaking, proposed outcomes and effect measures.
 
@@
 \begin{enumerate}
-  \item capture decision-relevant differences in patient benefit and harm
+  \item capture decision-relevant differences in distributions of
+    patient benefit and harm
   \item are easy to interpret and to translate into practical recommendations
@@
 
-\subsection{Tie-breaking can be used to fine-grain ordinal outcomes}
+\subsection{Tie-breaking can be used to combine and fine-grain ordinal outcomes}
 
-Two kinds: crossing, and tie breaking within a level (overriding).
+Two basic operations:
+\begin{enumerate}
+  \item Complete tie-breaking of $A$ by
+    $B$ (lexicographic order)
+  \item Tie-breaking of $A$ by $B$ \emph{within a
+    specific level $a$ of $A$} (equivalently, ``overriding'' $B$ by $A$) is the
+    combination of these.
+\end{enumerate}
+
+Tie breaking within a level (2) is a combination of complete
+tie-breaking (1) and merging.
+
+\subsubsection{Incorporating longitudinal or time-to-event information}
+
+Distinguish between ``state'' and ``trajectory'' variables. State
+variables summarize patient outcomes at a particular moment in time.
+Trajectory variables, including times-to-event, summarize patient
+state over a timecourse.
+
+Trajectory variables can be incorporated into an ordinal outcome. For example
+(from PASSPORT) ``worst pSOFA while hospitalized until day 28''.
+
+Incorporating trajectory variables into an ordinal outcome may seem unnatural.
+But arguable most existing time-to-event analyses are in fact analyses of an
+ordinal outcome. The Cox model for time-to-event data treats time as ordinal:
+because (1) the baseline hazard is not modelled, (2) the model itself is
+invariant to monotone rescalings of the time axis, and (3) the (partial)
+likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
+marginal likelihood). The Cox model is essentially an ordinal
+stopping ratio model with complementary log-log link. (Perhaps a
+difference is censoring.)
+
+\subsection{Coarsening (composite outcomes) can be used to avoid
+directly comparing levels}
+
+This comes from the experience of presenting the PASSPORT outcome:
+clinicians thought that including, for example, "digit amputation
+before day 28" and "ECMO before day 28" within the same level of an
+ordinal outcome meant assuming that the two outcomes have the
+\emph{same} utility. In fact it represents \emph{refusing to compare}
+their utilities.
 
@@
 
-\subsection{Ordinal outcomes can incorporate longitudinal or
-time-to-event information}
+Theory of \citep{} says that the relative efficiency (variance) of an
+unadjusted treatment comparison using an ordinal versus continuous outcome in
+the proportional odds model is given by $1 - \sum \overline p_c^3$ where
+$\overline p_c$ is the marginal probability (averaged over control and
+experimental groups) of outcome category $c$.
 
-Distinguish between ``state'' and ``trajectory'' variables.
+Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
+categories when the probabilities are all equal: $\overline p_c = 1/C$.
 
-Trajectory variables, including times-to-event, summarize over a timecourse.
+We have simulation results and some math to show that the value of $\sum_c
+\overline p_c^3$ is \emph{essentially a function of the largest category
+probability} $\overline{p_{\max}} = \max \overline{p_c}$. See \ref{fig:p_max}
+for an illustration.
 
-These can be incorporated into an ordinal outcome. For example ``worst pSOFA
-while hospitalized until day 28''.
+\begin{figure}
+  \includegraphics[width=6in]{p_max_controls_efficiency.png}
+  \caption{Bounds on the relative efficiency of an ordinal outcome
+    compared to its continuous counterpart as a function of the maximum
+  category probability.}
+  \label{fig:p_max}
+\end{figure}
 
-Arguably most existing time-to-event analyses are in fact analyses of an
-ordinal outcome. The Cox model treats time-to-event as an ordinal outcome:
-because (1) the baseline hazard is not modelled, (2) the model itself is
-invariant to monotone rescalings of the time axis, and (3) the (partial)
-likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
-marginal likelihood).
+Thus simple advice is: for statistical efficiency, try to make the proportion
+of patients in the largest category small.
 
@@
 
-\subsection{Model-based versus nonparametric effect measures}
+The basic problem: we want to compare two distributions (control and
+experimental groups) over ordinal outcomes with a single number.
 
-Model-based: common odds ratio; common stopping ratio or common
-continuation ratio.
+In general this represents a loss of information. Let's say the
+outcome has $L$ levels. Then the distributions in each
+arm is specified by $L-1$ numbers (for the first $L-1$ probabilities,
+  or cumulative logits, or stopping ratio logits--reflecting the
+sum-to-1 constraint on probability distributions) and so the
+\emph{difference} between the distributions is also specified by
+$L-1$ numbers (for example the differences between the cumulative
+logits, or logit stopping probabilities).
+
+An effect measure thus represents a way to collapse these $L-1$
+numbers into one.
+
+\subsection{Model-based effect measures}
+
+Model-based: common odds ratio; common stopping ratio or common continuation
+ratio.
+
+What they have in common: the effect measure is a byproduct of fitting a model
+to the data. The model typically assumes that a certain pattern holds in the
+outcome distributions across groups so that the differences in distributions
+can be described by a single number (with more than two groups, $G-1$ numbers).
+
+\subsection{Model-free effect measures}
 
 Nonparametric or model-free measures: probabilistic index, win ratio, win odds,
-...
+etc.
 
-Both of these can be considered ``estimands'', insofar as we consider.
+\subsubsection{Connection with win win/generalized pairwise comparison
+methods}
+
+Win/generalized pairwise comparison methods are implicitly about (1)
+constructing ordinal outcomes through tie-breaking (2) in a way that is
+compatible with right-censoring (3) estimating nonparametric effect
+measures (4) defining comparisons of two distributions by averaging
+over draws from the distributions as if they were independent
+(pairwise comparisons).
+
+\subsection{Comparing model-based and nonparametric effect measures}
+
+Both of these can be considered ``estimands'', insofar as we consider
+the model-based effect measures to be defined by the estimate from
+the best-fitting model in the population.
 
 Nonparametric effect measures may appear to be assumption-free. However
+
 \begin{itemize}
@@
     example there are nonmonotone effects across cutpoints
+  \item Model-free methods don't define a common scale for comparing
+    more than two interventions. For example, model-free effect
+    measures can be nontransitive
   \item Model-based methods have additional advantages, for example
@@
     personalized effect measure that is yet reportable as a single
-    number. Risk ratio for vaccine efficacy (being approximately
-    constant as a function of baseline risk/age) as an example.
+    number.
 \end{itemize}
@@
 
-\subsection{Generalized pairwise comparisons (``win'' methods)
-  are implicitly about constructing
-ordinal outcomes, in a way that is compatible with right-censoring}
+We should distinguish between effect measures used for
+\emph{analysis} and for \emph{reporting}. It may make sense to report
+the probabilistic index even if the analysis is undertaken using a
+proportional odds model.
+
+\section{Conclusions}
 

### 2025-05-13 notes_on_effect_measures.tex
Lots of work

\documentclass[
  11pt,
  fleqn
]{article}
% [fleqn] left-aligns all equations

\input{preamble/preamble}

% For writing/planning purposes: show paragraph level in TOC
\setcounter{secnumdepth}{2}

\addbibresource{bibliography.bib}
\graphicspath{{fig/}}

\title{Notes on Ordinal Measures of Association and Effect}
\author{Leon Di Stefano}
\date{\today}

\begin{document}

% \maketitle
\section*{Notes on Ordinal Measures of Association and Effect}
Leon Di Stefano

\section{Some definitions}

A good reference is \citet{kruskalOrdinalMeasuresAssociation1958}.

Two high level definitions/distinctions:

\begin{itemize}

  \item \textbf{``measures of association''
    versus ``measures of effect''}
    I want to distinguish measures of association, which measure the
    extent to which $p(x, y)
    \neq p(x)p(y)$, from ``measures of regression'' or ``measures of
    effect'' which measure differences among the conditional
    distributions $p(y \mid x)$ as $x$ varies, i.e.
    characterize the function $x \rightarrow p(y \mid x)$ (and in the
      simplest case when $x$ is binary, are comparisons of $p(y \mid
    x=1)$ to $p(y \mid x=0)$, as in treatment effects).

    How to assess whether a given measure is about ``association''
    versus ``effect''?

    In the strongest, fully nonparametric sense -- that is, without
    considering specific measures of effect or association -- ``absence of
    association'' and ``absence of effect'' coincide. But that
    doesn't help in characterizing what should count as a measure of
    association versus a measure of effect.

    Here are some ways to potentially formalize the difference:

    \begin{itemize}

      \item One way is to require that the formula for a measure
        of association should be symmetric in $x$ and $y$, whereas a
        measure of effect may not (or should not) be.

      \item Another way might be to require that measures of effect should be
        invariant to changing the marginal distribution of $x$ while leaving the
        conditionals intact (for example reweighting the joint
          distribution using
        weights that are a function only of $x$).

    \end{itemize}
    The second definition may be overly restrictive.

    Note that the odds ratio for a $2\times 2$ table satisfies both
    definitions above! It is a function of only the conditional
    distributions, but is also symmetric.

  \item \textbf{What does it mean to say that a measure is
    ``ordinal''?} Some measures of association or effect
    depend only on the probabilities attached to $x$ and
    $y$ values and not on any structure or properties of the $x$ and
    $y$ variables themselves. Call them ``distribution-only'' measures.

    An example is the mutual information for discrete $x, y$ given by
    \[ \sum_{x,
      y} p(x, y) \log\left(\frac{p(x, y)}{p(x) p(y)}\right) = \ave \left[
    \log\left(\frac{p(x, y)}{p(x) p(y)}\right) \right]. \]

    Others depend on the joint distribution of $x$ and $y$, but also
    only properties of $x$ and $y$ values themselves. Call these
    ``variable-involving'' measures.

    Covariance, for example, given by
    \[\cov(x, y) = \ave (x - \overline{x})(y - \overline{y}),\] is
    not like this,
    which depends on being able to add, subtract, scale, and
    multiply the associated variables.

    The importance of this definition is that when we consider
    ordinal measures of association and effect, we mean measures that
    are \emph{variable-involving, but through ordinal properties of
    the variables only}.

\end{itemize}

I'm not sure if there are more correct names for these in the literature.

\paragraph{Purely ordinal measures of association and effect} Ordinal measures
of effect should be functions of $y$ values but only through their ordering,
and ordinal measures of association should be functions of $x$ and
$y$ values, but only through their respective orderings. This means
they can make use of
\begin{itemize}
  \item Relations like $=, \leq, \geq$, which are functions of
    \emph{pairs} of values of a variable
  \item Quantiles (e.g. median) and order statistics, which are
    aspects of \emph{distribution} of an ordinal variable that depend
    only on the ordering; e.g. \[\text{median} = \text{solve}_q \left\{\Pr(x
    \leq q) = 0.5\right\}. \]
  \item Replacing $x$ and $y$ values by their ranks, or by the
    associated ``ridit'' (which is a rescaling of the ``midrank'',
    \citet{agrestiAnalysisOrdinalCategorical2010}) or eCDF
\end{itemize}

Another way of putting this
is that the measures should be
invariant under strictly monotone transformations of $x$ and $y$
\citep{kruskalOrdinalMeasuresAssociation1958}. However, this seems to
presume that the ordinal variables come from underlying numeric
variables, which may not be the case.

\paragraph{A 2$\times$2 breakdown} I think we have the options
\begin{itemize}
  \item $x$ ordinal, $y$ ordinal, measure of effect
  \item $x$ ordinal, $y$ ordinal, measure of association
  \item $x$ binary, $y$ ordinal, measure of effect
  \item $x$ binary, $y$ ordinal, measure of association
    (\textbf{requirement of symmetry means this is probably subsumed
    in the ordinal/ordinal case})
\end{itemize}
... which generalize the measures of effect and measures of
association for 2x2 tables.

\section{Constructing measures of association and effect using
independent draws}

The operators $=, <, >$ are
functions of \emph{pairs} of values. I think this is
one way to motivate the
following ``devices'' which can be used to construct ordinal measures of
association and
effect:
\begin{itemize}
  \item For measures of association, consider drawing
    \textbf{independent pairs} $z_1 = (x_1, y_1)$ and $z_2 = (x_2, y_2)$
    from $p(x, y)$.

    Then when $x$ and $y$ are ordinal, we can use $<, =, >$ to
    construct the disjoint events
    concordance $C(z_1, z_2)$, discordance $D(z_1, z_2)$, and ties
    in one or both
    variables $T_x(z_1, z_2), T_y(z_1, z_2), T_{xy}(z_1, z_2)$.

    We get measures of association by understanding the probabilities
    of these events under independence.

    Specifically,
    \begin{itemize}
      \item $\Pr(C) = \Pr(D)$ [I think]
      \item $\Pr(T_{xy}) = [\Pr(T_x) + \Pr(T_{xy})]\times [\Pr(T_y) +
        \Pr(T_{xy})]$
      \item The probabilities of $C, D, T_x, T_y, T_{xy}$ (5
          probabilities, 4 degrees
        of freedom) can be
        parametrized by only $T_x$, $T_y$. [I think]
    \end{itemize}

  \item For measures of effect when $x$ is binary, considering drawing
    \textbf{independent values of the response} $y_0
    \sim p(y \mid x = 0)$ and $y_1 \sim p(y \mid x = 1)$.

    Then when $y$ is ordinal we can consider the events $G = y_1 >
    y_0$ (``greater than''), $L = y_1 < y_0$ (``less than''), $E =
    y_1 == y_0$ (``equal to'').

    We get measures of effect by understanding the probabilities of these events
    under equality of distributions, or equivalently under
    independence of $y$ and binary $x$. Specifically, when the
    distributions are the same $\Pr(G) + \Pr(E)/2 = 1/2$
    and $\Pr(G) = \Pr(L)$.

\end{itemize}

\paragraph{Connection between the two devices}What's the connection
between the ``independent pairs'' and ``independent response values''
devices above? Consider $y$ ordinal
and $x$ \emph{dichotomous ordinal} with $x = 0, 1$ and $0 < 1$. (This
seems always true when we're considering an effect measure.) Then
\begin{itemize}
  \item The second sampling scheme is like the first but conditional
    on $x_1 \neq x_2$ [... I think].
  \item $C = G$ and $D = L$, and $T_y = E$. The events $T_x$ and
    $T_{xy}$ are excluded by construction.
\end{itemize}
That is, \textbf{every ordinal measure of association should
specialize to an ordinal measure of effect when $x$ is binary}
(though as discussed above \emph{maybe} we want this to be a true measure of
effect that \textbf{doesn't depend on the $x$-probabilities}).

Two asides:

\begin{itemize}
  \item \citet{kruskalOrdinalMeasuresAssociation1958} even
    considers further involving \emph{three} independent copies of
    the variables.
  \item Covariances and least squares regression coefficients can also be
    expressed using the ``pairs of draws'' device:\footnote{ I believe this is
      connected with
      $U$-statistics.\url{https://stats.stackexchange.com/questions/612496/linear-regression-vs-average-of-slopes}
    } \[ \cov(x, y) = \frac{1}{2}\ave(x - x')(y - y') \\ b_{yx} =
      \frac{\ave (y -
      y')(x - x')}{\ave (x - x')^2} = {\ave \left(w \times
      \text{pair-slopes}\right)}
    \]
    ... where the weight $w_{ij}$ is ${\Delta x^2}\big/{\sum \Delta x^2}$.

    As covered below some of the measures can be interpreted in terms
    of means and (co)variances and regression coefficients.
\end{itemize}

\paragraph{How this works in finite samples} When we have a dataset with $N$
values of $(x, y)$, then there are $N(N-1)$ \emph{ordered} pairs. If $x$ is
dichotomous with $n_0$ values of 0 and $n_1$ values of 1, \todo{...continue
this logic}.

\paragraph{Expression as regressions} Sometimes it's useful to
re-express these measures in terms of expectations, covariances,
correlations, etc. of e.g. ranks, or signs of differences in ranks, or

\begin{itemize}
  \item Spearman's $\rho$ = correlation of ranks
  \item Correlation of normal scores of ranks
\end{itemize}

\section{Graphical interpretations}

ROC curve: plots sensitivity against (1 - specificity). (Unscaled:
true positives against false positives.) That is, $\Pr(\text{correct}
\mid t, \text{False})$ against $\Pr(\text{incorrect} \mid t, \text{False})$..

P-P plot: plots eCDF(y) against eCDF(x) (or CDF($x$) in the case of a
theoretical reference distribution). Note that $x$ and $y$ are
ordinal but do not necessarily share the same space.

I think that \textbf{choosing a point at random in $PP$-space
  corresponds to independently choosing $x$ and a $y$ from their
marginal distributions}. Yes: see Figures
\ref{fig:pp_effect_measures} and \ref{fig:pp_ridit}.

What's the connection? Let $y$ be the score associated with the
diagnostic test and assume that a smaller $y$ is associated with a
decision in favor of $X =$ True.

Then the ROC curve is a plot of $\Pr(y \leq t \mid X = \text{True})$
(which is just sensitivity)
versus $\Pr(y \leq t \mid X = \text{False})$ (which is just 1 - specificity).

\begin{figure}
  \includegraphics[width=5in]{PP_or_ridit_interpretation_of_AUROC.png}
  \caption{PP-plot/ridit interpretation of AUROC, from
  \citet{smithsonReceiverOperatingCharacteristic2023}}
  \label{fig:pp_ridit}
\end{figure}

\begin{figure}
  \includegraphics[width=6.5in]{effect_measures_pp_plot.pdf}
  \caption{PP-plot/ridit interpretation of other effect measures,
  based on Figure \ref{fig:pp_ridit}}
  \label{fig:pp_effect_measures}
\end{figure}

\begin{figure}
  \includegraphics[width=6.5in]{ordinal_xy_pair_signs_plot.pdf}
  \caption{Plot illustrating components of ordinal-ordinal measures
    of association and effect. Regions of the same color in this plot
    and in the PP plot Figure \ref{fig:pp_effect_measures} correspond
    when $x$ is binary. Note that $x^*$ and $y^*$ have symmetric
  distributions centered at 0 by construction.}
  \label{fig:ordinal_xy}
\end{figure}

\section{List of measures of association and effect}

\subsection{Measures of effect}

\subsubsection{Binary $x$, ordinal $y$}

\begin{itemize}
  \item \textbf{Mann-Whitney $U$ statistic and its limit} Rescaling
    gives estimate of ``common language effect size'' = probabilistic
    index = probability of superiority = AUROC = win probability.... See
    \url{https://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test#Common_language_effect_size}.
      Note that this page also mentions a statistic ``$\rho$''
      which is the version incorporating ties, i.e.~$\Pr(Y > X) +
      0.5 \times \Pr(Y = X)$.
    \item \textbf{Wilcoxon statistic and its limit} Signed rank test. Either a
      single sample (in which case testing "whether the data come from
      a symmetric distribution with a specified center") or paired samples.
    \item \textbf{Win probability} Measure of effect. In terms of
      Somers' D it's
      $\frac{\text{SD} + 1}{2}$
    \item \textbf{Win difference} Seems to be the specialization of
      Kendall's Tau (see below).
    \item \textbf{Win ratio} Pr(win)/Pr(loss) (either matched or unmatched)
    \item \textbf{Win odds} Differs from the win ratio by incorporating ties
    \item \textbf{Mean ridit}
  \end{itemize}

  \subsubsection{Ordinal $x$, ordinal $y$}

  \begin{itemize}
    \item \textbf{Somers' $D$} \citep{newsonInterpretationSomersFour,
      newsonParametersNonparametricStatistics2002} Somers $D$ $=$ win
      difference (when $X$ is binary) = risk difference (when $X$ and
      $Y$ are binary)
    \item \textbf{Harrell's $C$ index} Equal to ROC with a binary outcome. A
      transform of Somers' $D$
    \item \text{Concordance/discordance ratio} = analogue of win
      ratio for ordinal-ordinal association
      \citet{newsonParametersNonparametricStatistics2002}. Reduces to
      odds ratio when $x$ and $y$ are binary.
  \end{itemize}

  Are there others in \citet{agrestiAnalysisOrdinalCategorical2010}
  that I'm missing?

  \subsection{Measures of association}

  \subsubsection{Ordinal $x$, ordinal $y$}

  \begin{itemize}
    \item \textbf{Kendall's $\tau$ ($a$ and $b$ and $c$ forms --
      \textbf{distinguish these})}
      From \citet{newsonParametersNonparametricStatistics2002}
      \begin{align*}
        \tau_a(X,Y)
        &= \mathrm{E}\left(\text{sign}(X_1-X_2)\text{sign}(Y_1-Y_2)\right) \\
        &=
        \mathrm{Cov}(\left(\text{sign}(X_1-X_2)\text{sign}(Y_1-Y_2)\right)) \\
        &= \Pr\left(\text{concordant}\right) -
        \Pr\left(\text{discordant}\right) = \Pr\left(C\right) -
        \Pr\left(D\right)
      \end{align*}
      Note that ``$x_1 - x_2$'' has no meaning for ordinal variables, but
      sign$(x_1 - x_2)$ is a meaningful abuse of notation.

      Somers' $D = D_{yx}$ is the associated regression coefficient
      $\tau_a(X, Y)/\tau_a(X, X)$

      Kendall's $\tau_b$ is the associated \emph{correlation} coefficient
      and is given by
      \[
        \tau_b = \text{sign}(\tau_a) \times \sqrt{D_{yx} D_{xy}}
      \]

    \item \textbf{Goodman-Kruskal $\gamma$} \[\gamma :=\frac{\Pr(C) -
      \Pr(D)}{\Pr(C) + \Pr(D)}\] Wikipedia: specializes to Yule's $Q$
      for 2x2 tables.

  \end{itemize}

  \newpage

  \section{References}
  \printbibliography

  \end{document}

### 2025-05-13 paper.tex
Lots of work

--- 2025-04-22 paper.tex
@@
 % For writing/planning purposes: show paragraph level in TOC
-\setcounter{secnumdepth}{2}
+\setcounter{secnumdepth}{3}
 
@@
 
-\title{Guidance on choosing ordinal outcomes and endpoints for clinical trials}
-\author{Leon Di Stefano, Katherine Lee, etc\ldots}
+\title{Guidance for clinicians in constructing ordinal outcomes and choosing
+ordinal effect measures for clinical trials}
+\author{Leon Di Stefano + Ordinal
+Working Group}
 \date{\today}
@@
 Ordinal outcomes have become increasingly prominent in clinical
-resarch, particularly following the COVID-19 pandemic.
+resarch, particularly following the COVID-19 pandemic, when ordinal
+scales developed by the WHO and NIAID were used in prominent trials
+of drugs repurposed to treat the disease \todo{cite}. They have also
+been a popular choice in adaptive platform trials \todo{cite}.
 
@@
   \item Provide more information that binary outcomes, including
-    composite outcomes, allowing for more efficient and ethical studies
-  \item Allow for the incorporation of safety events into one's
-    primary outcome and therefore account for many different aspects
-    of benefit/harm. At the same time, they do not require a
-    fully-specified utility function (which may differ from patient
-    to patient or require more extensive development and validation)
+    composite outcomes, making study designs more sensitive and precise, and
+    therefore more efficient and ethical
+
+  \item Allow for the incorporation of intercurrent events into one's primary
+    outcome, avoiding complexities of alternative estimand strategies like
+    principal stratum estimands (particular for truncating events
+    like death) while
+    yielding more statistical efficiency than traditional binary
+    composite outcomes
+
+  \item Allow for the easy incorporation of safety events into a
+    primary outcome, accounting for many different aspects
+    of benefit/harm; at the same time, ordinal outcomes can be
+    analyzed without a
+    fully-specified utility function, which may require more
+    extensive development and validation or may vary widely across patients.
+
   \item Allow for the combination of discrete and continous outcome
-    measures (for example a numerical quality of life scale with
-    death as the worst outcome)
-  \item Allow for the incorporation of intercurrent events into one's
-    primary outcome, avoiding complexities of alternative estimand strategies
-    like principal stratum estimands (particular for truncating
-    events like death) while having more statistical efficiency than
-    binary composite estimands (see point (1))
+    measures (for
+      example a numerical quality of life scale with death as the
+      overriding worst
+    outcome) and allow for the combination of longitudinal and timepoint or
+    ``state'' outcomes in a single measure.
+
 \end{enumerate}
@@
 The goal of this manuscript is to serve as a guide for clinicians and trialists
-in choosing/developing ordinal outcomes for their studies. We consider both
-both the outcome itself (the data recorded on each patient) and effect
-measures or ``endpoints'' (the metric used to compare patient groups).
+in developing ordinal outcomes for their studies. We consider both the outcome
+itself (the data recorded on each patient) and effect measures or ``endpoints''
+(the metric used to compare distributions of the outcome across patient
+groups).
 
-We want to distinguish \emph{choosing} an outcome/effect
-measure from \emph{validating} the outcome/effect measure using
-external data. This paper is about choosing the outcome and effect
-measure. In practice choosing an outcome and effect measure and
-validating that outcome and effect measure in pre-trial data will be
-an iterative process.
+We want to distinguish \emph{choosing} or \emph{constructing} an outcome/effect
+measure from \emph{validating} the outcome/effect measure using external data.
+This paper is about the first task, choosing (or constructing) the outcome and
+effect measure. In practice choosing an outcome and effect measure and
+validating that outcome and effect measure in pre-trial data will be an
+iterative process. Our focus in this paper is on the ``positive'' side of the
+process--namely coming up with, and tweaking, proposed outcomes and effect
+measures--rather than the ``negative'' side, namely testing or checking
+proposals with data to assess whether they have desired properties.
+
+A final goal of this paper is to situate the literature around ``generalized
+pairwise comparisons of prioritized outcomes''
+\citep{buyseGeneralizedPairwiseComparisons2022}--including win ratio
+\citep{pocockWinRatioNew2012} and Desirability of Outcome Ranking (DOOR)
+methods \citep{evansDesirabilityOutcomeRanking2015,
+ongUnlockingDOORHow2023}--within the broader context of ordinal outcomes and
+endpoints.
 
@@
 
+We begin with some general desiderata for outcomes and effect measures in
+clinical studies.
+
 We want outcomes that
 \begin{enumerate}
-  \item reflect patient benefit and harm
-  \item capture sufficient information to efficiently discriminate
-    among better- and
-    worse-off patients.
+  \item are \textbf{valid} in that they reflect patient benefit and harm:
+    better/worse outcomes are judged respectively better/worse by patients, are
+    objectively respectively better/worse for patients, or are reliable
+    surrogates/predictors of outcomes that are respectively better/worse for
+    patients
+  \item are \textbf{efficient} in the sense of capturing
+    sufficient information to
+    discriminate among better- and worse-off patients without large
+    sample sizes.
+\end{enumerate}
+(For the purposes of this article, set aside ease of collection, low
+missingess, objective ascertainability, etc.)
+
+An example of an outcome which is valid but inefficient might be
+death as a binary outcome in a context where deaths
+are rare.
+
+An example of an outcome which is precise but which may be invalid
+might be continuously-measured concentrations of a biomarker which is a poor
+surrogate for disease progression.
+
+What about effect measures? We generally seek effect measures that
+\begin{enumerate}
+  \item accurately summarize decision-relevant differences in distributions of
+    patient benefit and harm
+  \item are easy to interpret and to translate into practical recommendations
+  \item can be incorporated into statistical models where required
+  \item are easy to generalize or ``transport'' across contexts or settings
 \end{enumerate}
 
-We want effect measures that
+\section{Constructing and combining ordinal outcomes}
+
+\subsection{Fine-graining using tie-breaking and overrides for efficiency}
+
+Term in the literature: ``hierarchical composite endpoints''
+\citep{gasparyanDesignAnalysisStudies2022}.
+
+Two basic operations:
 \begin{enumerate}
-  \item capture decision-relevant differences in patient benefit and harm
-  \item are easy to interpret and to translate into practical recommendations
+  \item Complete tie-breaking of $A$ by
+    $B$ (``prioritized outcomes''; lexicographic order)
+  \item Tie-breaking by $B$
+    \emph{within specific levels of $A$}. This is equivalent to complete
+    tie-breaking followed by coarsening.
+  \item A special case of this is
+    \emph{overriding} of $B$ by $A$ when $A$ is binary.
 \end{enumerate}
 
-\section{Techniques for constructing ordinal outcomes}
+Examples, using "$/$" to denote overriding and "$\times$" to denote complete
+tiebreaking (note that this is an asymmetric operation):
+\begin{itemize}
+  \item
+    ``Angina Symptom Score'' outcome from
+    \citet{rajkumarPlaceboControlledTrialPercutaneous2023}, depicted in Figure
+    \ref{fig:rajkumar_et_al_outcome}, is given by \[ \text{death} / \text{ACS} /
+      \text{unacceptable angina} / (\text{antianginal meds} \times
+        \text{\# angina
+    episodes}) \]
+  \item Outcome from PASSPORT, depicted in Figure
+    \ref{fig:passport_outcome}, is given by \[ \text{death} / \text{SAE} /
+      ((\text{max pSOFA} > 1) \times (\text{if max pSOFA > 1: max
+            pSOFA else: days
+    hospitalized})) \]
+\end{itemize}
 
-\subsection{Tie-breaking can be used to fine-grain ordinal outcomes}
+Aside: \citet{wittkowskiCombiningSeveralOrdinal2004} consider combining
+ordinal variables into partial orders. This makes more sense in the
+generalized pairwise comparisons framework.
 
-Two kinds: crossing, and tie breaking within a level (overriding).
+\begin{figure}
+  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
+  \caption{``Angina Symptom Score'' ordinal outocome develped in
+    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
+    (1) complete tie-breaking (of units of antianginal medication by
+    episodes of angina) and (2) overriding (of angina episodes and
+      angina medication by adverse clinical events; of all other outcomes
+  by death; etc.)}
+  \label{fig:rajkumar_et_al_outcome}
+\end{figure}
 
-\subsection{Category probabilities affect precision and power mostly via the
-probability of the largest category}
+\begin{figure}
+  \includegraphics[width=5.5in]{passport_ordinal_outcome_tree.pdf}
+  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
+  \caption{Ordinal outcome used in the planned PASSPORT adaptive
+    platform trial. This illustrates (1) overriding (of hospital
+    outcomes by pSOFA outcomes, etc.) but also (2) using tie-breaking
+    to augment a ``state'' outcome (status at day 28) with
+    longitudinal (maximum pSOFA while hospitalized) and time-to-event
+    (days hospitalized)
+  information}
+  \label{fig:passport_outcome}
+\end{figure}
 
-\subsection{Ordinal outcomes can incorporate longitudinal or
-time-to-event information}
+\subsubsection{Tie breaking can augment a state outcome with longitudinal
+information}
 
-Distinguish between ``state'' and ``trajectory'' variables.
+Distinguish between ``state'' and ``trajectory'' variables. State
+variables summarize patient outcomes at a particular moment in time.
+Trajectory variables, including times-to-event, summarize patient
+state over a timecourse.
 
-Trajectory variables, including times-to-event, summarize over a timecourse.
+Trajectory and state variables can be incorporated into an ordinal outcome. For
+example (from PASSPORT) ``worst pSOFA while hospitalized until day 28'' and
+``length of hospitalization'', which are longitudinal variables (summary of a
+trajectory), are combined with the day 28 state variable ``death''.
 
-These can be incorporated into an ordinal outcome. For example ``worst pSOFA
-while hospitalized until day 28''.
-
-Arguably most existing time-to-event analyses are in fact analyses of an
-ordinal outcome. The Cox model treats time-to-event as an ordinal outcome:
+Incorporating trajectory variables into an ordinal outcome may seem unnatural.
+But arguably most time-to-event analyses are in fact analyses of an
+ordinal outcome. The Cox model for time-to-event data treats time as ordinal:
 because (1) the baseline hazard is not modelled, (2) the model itself is
@@
 likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
-marginal likelihood).
+marginal likelihood). The Cox model is essentially an ordinal
+stopping ratio model with complementary log-log link. (Perhaps a
+difference is censoring.)
 
-\section{Guidelines for choosing estimands or effect measures}
+\subsubsection{Overriding can augment a continuous outcome}
 
-\subsection{Model-based versus nonparametric effect measures}
+For example overriding a continuous outcome by death. It would be
+good to have an example of this.
 
-Model-based: common odds ratio; common stopping ratio or common
-continuation ratio.
+\subsubsection{Precision and power depend mostly on the probability of the
+largest category}
 
-Nonparametric or model-free measures: probabilistic index, win ratio, win odds,
-...
+Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
+efficiency (variance) of an unadjusted treatment comparison using an ordinal
+versus continuous outcome in the proportional odds model is given by $1 - \sum
+\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
+over control and experimental groups) of outcome category $c$.
 
-Both of these can be considered ``estimands'', insofar as we consider.
+Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
+categories when the probabilities are all equal: $\overline p_c = 1/C$.
+
+We have simulation results and some math to show that the value of $\sum_c
+\overline p_c^3$ is \emph{essentially a function of the largest category
+probability} $\overline{p_{\max}} = \max \overline{p_c}$. See \ref{fig:p_max}
+for an illustration.
+
+\begin{figure}
+  \includegraphics[width=6in]{p_max_controls_efficiency.png}
+  \caption{Bounds on the relative efficiency of an ordinal outcome
+    compared to its continuous counterpart as a function of the maximum
+    category probability $p_\text{max}$. The upper bound (best case
+    relative efficiency for a given $p_\text{max}$) is always achieved in the
+    limit of infinitely many categories ($C = \infty$). The lower bound
+    is always achieved when the number of categories is as small as
+  possible for a given $p_\text{max}$.}
+  \label{fig:p_max}
+\end{figure}
+
+Thus simple advice is: for statistical efficiency, try to make the proportion
+of patients in the largest category small.
+
+However a converse is: if the proportion of patients in the largest category is
+small, not much is gained by tie-breaking.
+
+\subsection{Coarsening and incorporating intercurrent events for validity}
+
+\subsubsection{Coarsening to avoid directly comparing levels}
+
+In the PASSPORT outcome (Figure \ref{fig:passport_outcome}), for
+example, we include both digit
+amputation and ECMO
+any time within 28 days post-enrollment under the heading of
+``adverse clinical events''.
+
+Including these outcomes within the same level of an
+ordinal outcome does not mean assuming that the two outcomes have the
+\emph{same} utility. Rather it represents \emph{avoiding comparing}
+their utilities. (This was something clinicians asked about more than once.)
+
+\subsubsection{Coarsening for simplicity}
+
+Too many levels can make it hard to directly compare meaningful
+proportions/probabilities.
+
+(It isn't really a problem for model fitting with modern software.)
+
+Coarsening, including combining levels and replacing complete
+tie-breaks with overrides, could help with this.
+
+\subsubsection{Incorporating intercurrent events using tie-breaking
+or overriding}
+
+The intercurrent event could be (1) incorporated as a level (2) used
+for tie-breaking or overriding.
+
+This is a version of the ``composite strategy'' as described in e.g.
+\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
+more efficiency than creating a \emph{binary} composite outcome.
+
+\section{Choosing ordinal effect measures}
+
+Goal: we want to compare two or more distributions over ordinal
+outcomes with a single number. (Two distributions when comparing control and
+experimental groups; more distributions when e.g. adjusting for covariates.)
+
+In general this represents a loss of information. Let's say the
+outcome has $C$ categories. Then the distribution in each
+arm is specified by $C-1$ numbers (for the first $C-1$ probabilities,
+  or cumulative logits, or stopping ratio logits--reflecting the
+sum-to-1 constraint on probability distributions) and so the
+\emph{difference} between the distributions is also specified by
+$C-1$ numbers (for example the differences between the cumulative
+logits, or logit stopping probabilities).
+
+An effect measure thus represents a way to collapse these $C-1$
+numbers into one.
+
+It usually doesn't make sense to consider effect measures based on
+numeric coding (e.g. difference in mean levels). This is adding
+information that is not present in the basic ordinal structure.
+
+\subsection{Model-based effect measures}
+
+Model-based: common odds ratio; common stopping ratio or common continuation
+ratio.
+
+What they have in common: the effect measure is a byproduct of fitting a model
+to the data.
+
+The model typically assumes that a certain pattern holds in the outcome
+distributions across groups. If the pattern holds, then the differences in
+distributions can be described by a single number (with more than two groups,
+$G-1$ numbers) instead of $(C-1)$ numbers.
+
+\subsection{Model-free or nonparametric effect measures}
+
+These are not based on a model (assumed pattern in the differences
+between outcome distributions).
+
+\subsubsection{Effect measures based on ``independent pairwise
+comparison'' construction}
+
+Probabilistic index/probability of
+superiority, win ratio, win odds, etc. These predate win ratio and DOOR methods
+\citep[e.g.][p.~14]{agrestiAnalysisOrdinalCategorical2010}.
+
+Rather they are all be based on the ``\emph{pairs construction}'',
+where we imagine independently sampling patients from the control and
+treatment groups and considering ``wins'' ``ties'' and ``losses''
+(corresponding to concordant and discordant pairs and ties in $y$
+respectively). They can
+also be \emph{estimated from data} using pairs of patients, but this
+is not essential and they could also be estimated by fitting a
+proportional odds model, for example (though they would not be a
+simple function of a coefficient).
+
+\subsubsection{Dichotomizing or comparing with a reference group}
+
+\begin{itemize}
+  \item Reducing to a binary outcome by picking a specific cutpoint,
+    e.g. ``hospitalization or
+    worse''. This is effectively using a binary outcome, but with the
+    ordinal structure used to gain power if fitting a model.
+  \item Reducing to a binary outcome using quantiles of a reference
+    group; for example,
+    ``worse than the median outcome under standard care''
+  \item Ridit analysis
+    \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
+    smithsonReceiverOperatingCharacteristic2023, jansenRiditAnalysisReview1984}
+\end{itemize}
+
+\subsection{Comparing model-based and nonparametric effect measures}
+
+Both of these can be considered ``estimands'', insofar as we consider the
+estimand associated with the model-based effect measures to be given by the
+best-fitting model in the population.
+
+Two classes of assumption violations for model-based effect measures
+are \emph{magnitude}-heterogeneity across cutpoints (for example,
+  effect of a treatment on hospital length of stay or worse is larger
+  than its effect on
+death on the relevant scale) and
+\emph{sign}-heterogeneity
+across cutpoints (failure of stochastic dominance -- for example,
+treatment is beneficial for hospital length of stay, harmful for death).
 
 Nonparametric effect measures may appear to be assumption-free. However
+
 \begin{itemize}
@@
     effect measures to make sense
-  \item The model-free methods also may not make sense, if for
-    example there are nonmonotone effects across cutpoints
-  \item Model-based methods have additional advantages, for example
+  \item In particular, if stochastic dominance holds, we can think of
+    both model-based and nonparametric effect measures as
+    testing the null hypothesis of no effect ``in a particular
+    direction''. A weak interpretation of any particular
+    effect measure
+    is that it supplies this ``direction''.
+
+    For example the Wilcoxon test is
+    equivalent to the score test from fitting a proportional odds model.
+    Thus fitting a proportional odds model (or any other model) has a
+    valid null-testing
+    interpretation even if proportional odds doesn't hold.
+  \item Under failure of stochastic dominance (``sign
+    heterogeneity''), \emph{neither} parametric nor nonparametric
+    effect measures make sense, since it doesn't make sense to
+    summarize effects by a single number. Need to dig into the
+    cutpoint-specific effects.
+  \item Model-free methods don't define a common scale for comparing
+    more than two interventions. For example, model-free effect
+    measures can be \emph{nontransitive}, so it's not clear how to
+    use them in $\geq 2$-arm studies
+  \item For the same reason, model-based methods have additional
+    advantages, for example
     the possibility of adjustment for covariates.
@@
     worlds'' effect measure that is
-    personalized effect measure that is yet reportable as a single
-    number. Risk ratio for vaccine efficacy (being approximately
-    constant as a function of baseline risk/age) as an example.
+    personalized (conditional) yet reportable as a single
+    number.
+\end{itemize}
+
+Harrell critiques \citep{harrellOverviewCompositeOutcome2024,
+harrellRareDegenerativeDiseases2024, harrellViewsCompositeOutcome}: mainly that
+win probability etc. are \emph{relative measures}:
+
+\begin{itemize}
+  \item Comparable to Cohen's $d$ or $t$-statistic (difference in
+    means divided by standard deviation or standard error) rather than
+    difference in means: compare: standardized effect size of 0.2 vs.
+    difference in means of 10 mmHg
+  \item Affected by study inclusion criteria (why? probably needs to
+    be fleshed out.)
 \end{itemize}
@@
 Simulations show a typically very strong association between
-proportional odds ratios and probabilistic indices [cite Harrell].
+proportional odds ratios and $c$-index
+\citep{harrellViolationProportionalOdds2020}.
 
-Reducing to a composite binary outcome
+One point of view is that one would ``ideally'' analyze ordinal
+outcomes by reporting the associated expected utilities \todo{cite Berry
+talk; example of utility-weighted Rankin scale} and that since
+existing ordinal methods supply such utilities implicitly why not
+just supply them explicitly. A counterpoint is that the methods do not, in fact,
+supply utilities implicitly.
+
+We should distinguish between effect measures used for \emph{analysis} and for
+\emph{reporting}. It may make sense to report the probabilistic index or a
+single-cutpoint binary outcome even if the analysis is undertaken using a
+proportional odds model.
+
+\section{Situating win/DOOR/etc. methods in the broader context of ordinal
+outcomes and effect measures}
+
+Terms to define/distinguish:
 \begin{itemize}
-  \item Based on a specific cutpoint, e.g. ``hospitalization or worse''
-  \item Based on a quantile in a reference group; for example,
-    ``worse than the median outcome under standard care''
+  \item Desirability of Outcome Ranking (outcome)
+  \item Hierarchical composite endpoint (outcome)
+  \item Prioritized outcomes (outcome)
+  \item Win methods (win probability, win odds, win ratio) (effect
+    measure; estimation methods)
+  \item Generalized pairwise comparisons (effect measure)
 \end{itemize}
 
-One point of view is that one would ``ideally'' analyze ordinal
-outcomes by reporting the associated expected utilities [cite Berry
-talk; example of utility-weighted Rankin scale].
+These approaches implicitly \textbf{bundle together three different things}:
+\begin{enumerate}
+  \item Constructing an implicit ordinal outcome through
+    complete tie-breaking
+  \item Targeting a nonparametric effect measure defined via the
+    ``pairs'' device (probabilistic
+    index/win ratio/win odds/etc.)
+  \item Estimating that effect measure in a nonparametric/flexible
+    manner using pairs, often also with devices for accounting for censoring
+\end{enumerate}
 
-\subsection{Generalized pairwise comparisons (``win'' methods)
-  are implicitly about constructing
-ordinal outcomes, in a way that is compatible with right-censoring}
+This paper is focused on first two steps.
+
+Why is it useful to consider these three steps separately?
+
+Consider some of the critiques of win methods in
+\citet[e.g.~][]{ajufoFallaciesUsingWin2023}
+
+\begin{itemize}
+  \item Situating methods in their proper context can help
+    trialists make informed choices among them
+  \item Win methods can make use of
+    plots and diagnostics for general ordinal outcomes, e.g.
+    comparative CDF plots.
+  \item Ordinal regression models can use win methods for model checking; it is
+    also possible to present a model-based win effect measure alongside a common
+    odds ratio for example.
+  \item The outcomes \emph{implicitly} constructed using
+    DOOR and win methods may contain many levels, and may contain
+    levels where an
+    effect is not so meaningful (e.g. time-to-death when the followup time is
+    short). By \emph{explicitly} constructing the associated ordinal outcome one
+    might be forced to be more aware of this
+  \item What is a reasonable win
+    analysis of a specific component score? From the traditional
+    ordinal endpoint
+    school of thought what would make most sense is to \emph{only
+      tie-break for the
+    lowest and highest levels of that score}, and e.g. plot eCDFs of this new
+    variable. Is this what the plots do that break down wins and losses by
+    component score?
+\end{itemize}
+
+Disadvantages of generalized pairwise comparison approaches
+\begin{itemize}
+  \item Inherited from downsides of complete tiebreaking
+  \item Inherited from downsides of nonparametric/pairwise effect measures
+  \item Regression methods exist, but are newer and less battle-tested
+\end{itemize}
+
+\todo{``Partial credit'' analysis?}
+
+\section{Conclusions}
+
+\newpage
+
+\section{Bibliography}
+
+\printbibliography
 

### 2025-05-15 paper.tex
Add BANDICOOT parallel composition example
Revert to simpler PASSPORT outcome diagram

@@
 \begin{figure}
-  \includegraphics[width=5.5in]{passport_ordinal_outcome_tree.pdf}
+  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
   % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
@@
   \label{fig:passport_outcome}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
+  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
+  \caption{Ordinal outcome used in the planned BANDICOOT adaptive
+    platform trial \citep{walkerCodesigningNovelOrdinal2025}.
+  Illustrates \textbf{parallel composition} of component outcomes}
+  \label{fig:parallel_bandicoot}
 \end{figure}

### 2025-05-15 notes_on_effect_measures.tex
Additional work

@@
       $U$-statistics.\url{https://stats.stackexchange.com/questions/612496/linear-regression-vs-average-of-slopes}
+      See also \url{https://www.fharrell.com/post/po/} discussion of
+      Mann-Whitney. This also says that \textbf{it's possible to
+        convert between Wilcoxon statistic, Somers' $D$, concordance
+      probability, \emph{and} the Spearman rank correlation.}
     } \[ \cov(x, y) = \frac{1}{2}\ave(x - x')(y - y') \\ b_{yx} =
@@
 
-Then the ROC curve is a plot of $\Pr(y \leq t \mid X = \text{True})$
+Then the ROC curve is a plot of $\Pr(Y \leq t \mid X = \text{True})$
 (which is just sensitivity)
-versus $\Pr(y \leq t \mid X = \text{False})$ (which is just 1 - specificity).
+versus $\Pr(Y \leq t \mid X = \text{False})$ (which is just 1 -
+specificity) -- that is, a plot of cdf($y \mid x = 1$) versus cdf($y
+\mid x = 0$)
 
@@
 
-\subsubsection{Binary $x$, ordinal $y$}
+\subsubsection{Binary \emph{x}, ordinal \emph{y}}
 
@@
       0.5 \times \Pr(Y = X)$.
-    \item \textbf{Wilcoxon statistic and its limit} Signed rank test. Either a
-      single sample (in which case testing "whether the data come from
-      a symmetric distribution with a specified center") or paired samples.
+      % \item \textbf{Wilcoxon statistic and its limit} Signed rank
+      % test. Either a
+      %   single sample (in which case testing "whether the data come from
+      %   a symmetric distribution with a specified center") or paired samples.
     \item \textbf{Win probability} Measure of effect. In terms of
@@
 
-  \subsubsection{Ordinal $x$, ordinal $y$}
+  \subsubsection{Ordinal \emph{x}, ordinal \emph{y}}
 

### 2025-05-15 paper.tex
Additional work

@@
 
+Also, we "chunk" longer hospital stays more aggressively to avoid
+small categories, for interpretability's sake.
+
 \subsubsection{Coarsening for simplicity}
@@
 \section{Choosing ordinal effect measures}
+
+Key references:
+\citet{agrestiMeasuresNominalOrdinalAssociation1981,agrestiOrdinalProbabilityEffect2017a,
+agrestiSimpleWaysInterpret2018a}
 
@@
   \item Win methods can make use of
-    plots and diagnostics for general ordinal outcomes, e.g.
-    comparative CDF plots.
+    plots and diagnostics for general ordinal outcomes, for the
+    \emph{implicit} ordinal variable
+    \begin{itemize}
+      \item
+        comparative CDF plots for the
+      \item ridit plots
+        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984}
+        relative to a reference group, which should allow one to
+        display both full eCDF information and also win statistics
+        (based on the interpretation of the mean ridit as a win probability)
+    \end{itemize}e.g.
   \item Ordinal regression models can use win methods for model checking; it is
     also possible to present a model-based win effect measure alongside a common
-    odds ratio for example.
+    odds ratio for example. See
+    \citep{agrestiOrdinalProbabilityEffect2017} for example, which
+    covers both latent variable and predicted probability versions.
   \item The outcomes \emph{implicitly} constructed using

### 2025-05-15 paper.tex
Work before meeting presentation

@@
 
-\section{Desiderata for outcomes and effect measures}
+\section{Desiderata for outcomes and effect measures: validity and efficiency}
 
@@
 
-\section{Constructing and combining ordinal outcomes}
+\section{Combining and coarsening ordinal outcomes}
 
-\subsection{Fine-graining using tie-breaking and overrides for efficiency}
+\subsection{3 ways to combine/fine-grain ordinal outcomes: sequential,
+hierarchical, \& parallel composition}
 
@@
 
-Two basic operations:
+Three broad ways to combine ordinal outcomes:
 \begin{enumerate}
-  \item Complete tie-breaking of $A$ by
-    $B$ (``prioritized outcomes''; lexicographic order)
-  \item Tie-breaking by $B$
-    \emph{within specific levels of $A$}. This is equivalent to complete
-    tie-breaking followed by coarsening.
-  \item A special case of this is
-    \emph{overriding} of $B$ by $A$ when $A$ is binary.
+  \item \textbf{Sequential composition} Tie-breaking by $B$
+    \emph{within either highest or lowest level of $A$}. Also could
+    call this ``overriding by $A$''. A special case of this is
+    \emph{overriding events} when $A$ is binary.
+  \item \textbf{Hierarchical composition} Complete tie-breaking of $A$ by
+    $B$ (``prioritized outcomes''; lexicographic order) within
+    \emph{every} level of $A$.
+
+    This is what win methods do.
+  \item \textbf{Parallel composition} This is where levels of $A$ are
+    combined with levels of $B$ ``in parallel''; high levels of $A$
+    are combined with high levels of $B$ and low levels of $A$ with
+    low levels of $B$.
+
+    This is not so precisely determined as the other two. We need to
+    decided \emph{how} to aggregate.
 \end{enumerate}
 
-Examples, using "$/$" to denote overriding and "$\times$" to denote complete
-tiebreaking (note that this is an asymmetric operation):
+Suggested notation:
 \begin{itemize}
+  \item "$\rightarrow$" and "$\leftarrow$" to denote sequential
+    composition/overriding,
+  \item "$/$" to denote hierarchical composition/complete tiebreaking
+    (note that this is an asymmetric operation), and
+  \item "$|$" to denote parallel composition
+\end{itemize}
+
+Examples:
+
+\begin{itemize}
+  \item \textbf{Sequential composition} Outcome from PASSPORT,
+    depicted in Figure
+    \ref{fig:passport_outcome}, is given by
+    \[ \text{day 28 status} \leftarrow \text{max pSOFA} \leftarrow \text{days
+      hospitalized}
+    \]
   \item
-    ``Angina Symptom Score'' outcome from
+    \textbf{Hierarchical composition} ``Angina Symptom Score'' outcome from
     \citet{rajkumarPlaceboControlledTrialPercutaneous2023}, depicted in Figure
-    \ref{fig:rajkumar_et_al_outcome}, is given by \[ \text{death} / \text{ACS} /
-      \text{unacceptable angina} / (\text{antianginal meds} \times
-        \text{\# angina
-    episodes}) \]
-  \item Outcome from PASSPORT, depicted in Figure
-    \ref{fig:passport_outcome}, is given by \[ \text{death} / \text{SAE} /
-      ((\text{max pSOFA} > 1) \times (\text{if max pSOFA > 1: max
-            pSOFA else: days
-    hospitalized})) \]
+    \ref{fig:rajkumar_et_al_outcome}, is given by \[
+      \text{adverse clinical event} \leftarrow ((\text{units of
+    antianginal medication}) / (\text{episodes of angina})) \]
+
+  \item \textbf{Parallel composition} BANDICOOT outcome
+    \citep{walkerCodesigningNovelOrdinal2025}, which is
+    \[
+      (\text{death, graft loss, or relapse}) \leftarrow (\text{viraemia} \;|\;
+      \text{CD4+ count} \;|\; \text{organ support})
+    \]
 \end{itemize}
@@
 generalized pairwise comparisons framework.
+
+\subsubsection{Another way: scores}
+
+Treat each input factor as
+numeric then sum them up. I believe \citet{ongUnlockingDOORHow2023}
+has examples?
+
+\begin{figure}
+  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
+  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
+  \caption{Ordinal outcome used in the planned PASSPORT adaptive
+    platform trial. This illustrates sequential composition/overriding:
+    of days hospitalized by maximum pSOFA while hospitalized, and of
+  both of these by day 28 status.}
+  \label{fig:passport_outcome}
+\end{figure}
 
@@
     episodes of angina) and (2) overriding (of angina episodes and
-      angina medication by adverse clinical events; of all other outcomes
-  by death; etc.)}
+  angina medication by adverse clinical events)}
   \label{fig:rajkumar_et_al_outcome}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
-  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
-  \caption{Ordinal outcome used in the planned PASSPORT adaptive
-    platform trial. This illustrates (1) overriding (of hospital
-    outcomes by pSOFA outcomes, etc.) but also (2) using tie-breaking
-    to augment a ``state'' outcome (status at day 28) with
-    longitudinal (maximum pSOFA while hospitalized) and time-to-event
-    (days hospitalized)
-  information}
-  \label{fig:passport_outcome}
 \end{figure}
@@
 
-\subsubsection{Tie breaking can augment a state outcome with longitudinal
+\subsection{Reasons to combine ordinal outcomes}
+
+\subsubsection{Incorporating intercurrent events for validity}
+
+The intercurrent event could be (1) incorporated as a level (2) used
+for tie-breaking or overriding.
+
+This is a version of the ``composite strategy'' as described in e.g.
+\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
+more efficiency than creating a \emph{binary} composite outcome.
+
+For example overriding a numeric outcome by death.
+Example: REMAP-CAP: organ-free survival, overridden by death
+(Figure \ref{fig:remap_cap_override}).
+
+\begin{figure}
+  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
+  \caption{Ordinal outcome used in the REMAP-CAP adaptive
+    platform trial
+    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
+    Illustrates overriding a numeric ordinal outcome (organ
+    support-free days) by a binary ordinal outcome (death; this is
+    ``sequential composition''), which
+  would also be an intercurrent event.}
+  \label{fig:remap_cap_override}
+\end{figure}
+
+\subsubsection{Increasing precision by reducing the probability of the
+largest category}
+
+Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
+efficiency (variance) of an unadjusted treatment comparison using an ordinal
+versus continuous outcome in the proportional odds model is given by $1 - \sum
+\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
+over control and experimental groups) of outcome category $c$.
+
+Same should apply to the Mann-Whitney-Wilcoxon test, and therefore to
+the win probability, and therefore at least approximately to win
+difference/win ratio/win odds.
+
+Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
+categories when the probabilities are all equal: $\overline p_c = 1/C$.
+
+We have simulation results and some math to show that the value of $\sum_c
+\overline p_c^3$ is \emph{essentially a function of the largest category
+probability} $\overline{p_{\max}} = \max \overline{p_c}$. See Figure
+\ref{fig:p_max} for an illustration.
+
+\begin{figure}
+  \includegraphics[width=6in]{p_max_controls_efficiency.png}
+  \caption{Bounds on the relative efficiency of an ordinal outcome
+    compared to its continuous counterpart as a function of the maximum
+    category probability $p_\text{max}$. The upper bound (best case
+    relative efficiency for a given $p_\text{max}$) is always achieved in the
+    limit of infinitely many categories ($C = \infty$). The lower bound
+    is always achieved when the number of categories is as small as
+  possible for a given $p_\text{max}$.}
+  \label{fig:p_max}
+\end{figure}
+
+Thus simple advice is: for statistical efficiency, try to make the proportion
+of patients in the largest category small.
+
+However a converse is: if the proportion of patients in the largest
+category is
+small, not much is gained by tie-breaking.
+
+\subsubsection{Augmenting a state outcome with longitudinal
 information}
@@
 
-\subsubsection{Overriding can augment a continuous outcome}
-
-For example overriding a continuous outcome by death. It would be
-good to have an example of this.
-
-\subsubsection{Precision and power depend mostly on the probability of the
-largest category}
-
-Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
-efficiency (variance) of an unadjusted treatment comparison using an ordinal
-versus continuous outcome in the proportional odds model is given by $1 - \sum
-\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
-over control and experimental groups) of outcome category $c$.
-
-Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
-categories when the probabilities are all equal: $\overline p_c = 1/C$.
-
-We have simulation results and some math to show that the value of $\sum_c
-\overline p_c^3$ is \emph{essentially a function of the largest category
-probability} $\overline{p_{\max}} = \max \overline{p_c}$. See \ref{fig:p_max}
-for an illustration.
-
-\begin{figure}
-  \includegraphics[width=6in]{p_max_controls_efficiency.png}
-  \caption{Bounds on the relative efficiency of an ordinal outcome
-    compared to its continuous counterpart as a function of the maximum
-    category probability $p_\text{max}$. The upper bound (best case
-    relative efficiency for a given $p_\text{max}$) is always achieved in the
-    limit of infinitely many categories ($C = \infty$). The lower bound
-    is always achieved when the number of categories is as small as
-  possible for a given $p_\text{max}$.}
-  \label{fig:p_max}
-\end{figure}
-
-Thus simple advice is: for statistical efficiency, try to make the proportion
-of patients in the largest category small.
-
-However a converse is: if the proportion of patients in the largest category is
-small, not much is gained by tie-breaking.
-
-\subsection{Coarsening and incorporating intercurrent events for validity}
+\subsection{Reasons to coarsen ordinal outcomes}
 
@@
 
-\subsubsection{Coarsening for simplicity}
+\subsubsection{Coarsening for ease of communication}
 
@@
 tie-breaks with overrides, could help with this.
-
-\subsubsection{Incorporating intercurrent events using tie-breaking
-or overriding}
-
-The intercurrent event could be (1) incorporated as a level (2) used
-for tie-breaking or overriding.
-
-This is a version of the ``composite strategy'' as described in e.g.
-\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
-more efficiency than creating a \emph{binary} composite outcome.
 
@@
 
+\begin{figure}
+  \includegraphics[width=7in]{illustrating_common_or.pdf}
+  \caption{The common odds ratio, a model-based effect measure, comes
+    from the best-fitting parallel curves to the logit transformed
+    empirical cumulative distribution. \textbf{Ideally would illustrate
+      using toy bar charts as above, with only 2 groups? Although it's a
+  strength that model-based measures can compare multiple groups.}}
+  \label{fig:illustrating_or}
+\end{figure}
+
 Model-based: common odds ratio; common stopping ratio or common continuation
@@
 What they have in common: the effect measure is a byproduct of fitting a model
-to the data.
+to the data. See Figure \ref{fig:illustrating_or} for an illustration
+for the common odds ratio.
 
@@
 $G-1$ numbers) instead of $(C-1)$ numbers.
+
+In Figure \ref{fig:illustrating_or} the pattern assumed to hold is
+that the logit-transformed eCDFs in each group are parallel.
 
@@
 
+\begin{figure}
+  \includegraphics[width=7.5in]{effect_measures_pp_plot.pdf}
+  \caption{Illustrating independent pairwise comparison-based effect
+    measures using a PP-plot or ROC curve. Adapted from
+  \citet{smithsonReceiverOperatingCharacteristic2023}.}
+  \label{fig:illustrating_pairwise_comparisons}
+\end{figure}
+
 Probabilistic index/probability of
-superiority, win ratio, win odds, etc. These predate win ratio and DOOR methods
+superiority, win ratio, win odds, etc. These predate win ratio and
+DOOR methods
 \citep[e.g.][p.~14]{agrestiAnalysisOrdinalCategorical2010}.
@@
 (corresponding to concordant and discordant pairs and ties in $y$
-respectively). They can
+respectively). See Figure \ref{fig:illustrating_pairwise_comparisons}
+for an illustration.
+
+They can
 also be \emph{estimated from data} using pairs of patients, but this
@@
 \begin{itemize}
-  \item Reducing to a binary outcome by picking a specific cutpoint,
+  \item Dichotomizing by picking a specific cutpoint,
     e.g. ``hospitalization or
@@
     ordinal structure used to gain power if fitting a model.
-  \item Reducing to a binary outcome using quantiles of a reference
+  \item Dichotomizing using quantiles of a reference
     group; for example,
     ``worse than the median outcome under standard care''
-  \item Ridit analysis
+  \item Simultaneously comparing all quantiles to those of a
+    reference group: ridit analysis
     \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
-    smithsonReceiverOperatingCharacteristic2023, jansenRiditAnalysisReview1984}
+      smithsonReceiverOperatingCharacteristic2023,
+    jansenRiditAnalysisReview1984}
 \end{itemize}
@@
 Harrell critiques \citep{harrellOverviewCompositeOutcome2024,
-harrellRareDegenerativeDiseases2024, harrellViewsCompositeOutcome}: mainly that
+harrellRareDegenerativeDiseases2024, harrellViewsCompositeOutcome}:
+mainly that
 win probability etc. are \emph{relative measures}:
@@
 existing ordinal methods supply such utilities implicitly why not
-just supply them explicitly. A counterpoint is that the methods do not, in fact,
+just supply them explicitly. A counterpoint is that the methods do
+not, in fact,
 supply utilities implicitly.
@@
   \item Constructing an implicit ordinal outcome through
-    complete tie-breaking
+    hierarchical composition (complete tie-breaking/prioritized outcomes)
   \item Targeting a nonparametric effect measure defined via the
-    ``pairs'' device (probabilistic
-    index/win ratio/win odds/etc.)
+    ``pairs'' device (win probability/win ratio/win odds/etc.)
   \item Estimating that effect measure in a nonparametric/flexible
@@
     \end{itemize}e.g.
-  \item Ordinal regression models can use win methods for model checking; it is
-    also possible to present a model-based win effect measure alongside a common
+  \item Ordinal regression models can use win methods for model
+    checking; it is
+    also possible to present a model-based win effect measure
+    alongside a common
     odds ratio for example. See
@@
     effect is not so meaningful (e.g. time-to-death when the followup time is
-    short). By \emph{explicitly} constructing the associated ordinal outcome one
+    short). By \emph{explicitly} constructing the associated
+    ordinal outcome one
     might be forced to be more aware of this

### 2025-05-29 choosing_ordinal_effect_measures.tex
Rename + start effect measures draft

--- 2025-05-15 paper.tex
@@
 
-\title{Guidance for clinicians in constructing ordinal outcomes and choosing
-ordinal effect measures for clinical trials}
+\title{Choosing ordinal effect measures for clinical trials}
 \author{Leon Di Stefano + Ordinal
@@
 
-Ordinal outcomes have become increasingly prominent in clinical
-resarch, particularly following the COVID-19 pandemic, when ordinal
-scales developed by the WHO and NIAID were used in prominent trials
-of drugs repurposed to treat the disease \todo{cite}. They have also
-been a popular choice in adaptive platform trials \todo{cite}.
-
-Some advantages of ordinal outcomes are that they can
-\begin{enumerate}
-  \item Provide more information that binary outcomes, including
-    composite outcomes, making study designs more sensitive and precise, and
-    therefore more efficient and ethical
-
-  \item Allow for the incorporation of intercurrent events into one's primary
-    outcome, avoiding complexities of alternative estimand strategies like
-    principal stratum estimands (particular for truncating events
-    like death) while
-    yielding more statistical efficiency than traditional binary
-    composite outcomes
-
-  \item Allow for the easy incorporation of safety events into a
-    primary outcome, accounting for many different aspects
-    of benefit/harm; at the same time, ordinal outcomes can be
-    analyzed without a
-    fully-specified utility function, which may require more
-    extensive development and validation or may vary widely across patients.
-
-  \item Allow for the combination of discrete and continous outcome
-    measures (for
-      example a numerical quality of life scale with death as the
-      overriding worst
-    outcome) and allow for the combination of longitudinal and timepoint or
-    ``state'' outcomes in a single measure.
-
-\end{enumerate}
-
-The goal of this manuscript is to serve as a guide for clinicians and trialists
-in developing ordinal outcomes for their studies. We consider both the outcome
-itself (the data recorded on each patient) and effect measures or ``endpoints''
-(the metric used to compare distributions of the outcome across patient
-groups).
-
-We want to distinguish \emph{choosing} or \emph{constructing} an outcome/effect
-measure from \emph{validating} the outcome/effect measure using external data.
-This paper is about the first task, choosing (or constructing) the outcome and
-effect measure. In practice choosing an outcome and effect measure and
-validating that outcome and effect measure in pre-trial data will be an
-iterative process. Our focus in this paper is on the ``positive'' side of the
-process--namely coming up with, and tweaking, proposed outcomes and effect
-measures--rather than the ``negative'' side, namely testing or checking
-proposals with data to assess whether they have desired properties.
-
-A final goal of this paper is to situate the literature around ``generalized
-pairwise comparisons of prioritized outcomes''
-\citep{buyseGeneralizedPairwiseComparisons2022}--including win ratio
-\citep{pocockWinRatioNew2012} and Desirability of Outcome Ranking (DOOR)
-methods \citep{evansDesirabilityOutcomeRanking2015,
-ongUnlockingDOORHow2023}--within the broader context of ordinal outcomes and
-endpoints.
-
-\section{Desiderata for outcomes and effect measures}
-
-We begin with some general desiderata for outcomes and effect measures in
-clinical studies.
-
-We want outcomes that
-\begin{enumerate}
-  \item are \textbf{valid} in that they reflect patient benefit and harm:
-    better/worse outcomes are judged respectively better/worse by patients, are
-    objectively respectively better/worse for patients, or are reliable
-    surrogates/predictors of outcomes that are respectively better/worse for
-    patients
-  \item are \textbf{efficient} in the sense of capturing
-    sufficient information to
-    discriminate among better- and worse-off patients without large
-    sample sizes.
-\end{enumerate}
-(For the purposes of this article, set aside ease of collection, low
-missingess, objective ascertainability, etc.)
-
-An example of an outcome which is valid but inefficient might be
-death as a binary outcome in a context where deaths
-are rare.
-
-An example of an outcome which is precise but which may be invalid
-might be continuously-measured concentrations of a biomarker which is a poor
-surrogate for disease progression.
-
-What about effect measures? We generally seek effect measures that
-\begin{enumerate}
-  \item accurately summarize decision-relevant differences in distributions of
-    patient benefit and harm
-  \item are easy to interpret and to translate into practical recommendations
-  \item can be incorporated into statistical models where required
-  \item are easy to generalize or ``transport'' across contexts or settings
-\end{enumerate}
-
-\section{Constructing and combining ordinal outcomes}
-
-\subsection{Fine-graining using tie-breaking and overrides for efficiency}
-
-Term in the literature: ``hierarchical composite endpoints''
-\citep{gasparyanDesignAnalysisStudies2022}.
-
-Two basic operations:
-\begin{enumerate}
-  \item Complete tie-breaking of $A$ by
-    $B$ (``prioritized outcomes''; lexicographic order)
-  \item Tie-breaking by $B$
-    \emph{within specific levels of $A$}. This is equivalent to complete
-    tie-breaking followed by coarsening.
-  \item A special case of this is
-    \emph{overriding} of $B$ by $A$ when $A$ is binary.
-\end{enumerate}
-
-Examples, using "$/$" to denote overriding and "$\times$" to denote complete
-tiebreaking (note that this is an asymmetric operation):
-\begin{itemize}
-  \item
-    ``Angina Symptom Score'' outcome from
-    \citet{rajkumarPlaceboControlledTrialPercutaneous2023}, depicted in Figure
-    \ref{fig:rajkumar_et_al_outcome}, is given by \[ \text{death} / \text{ACS} /
-      \text{unacceptable angina} / (\text{antianginal meds} \times
-        \text{\# angina
-    episodes}) \]
-  \item Outcome from PASSPORT, depicted in Figure
-    \ref{fig:passport_outcome}, is given by \[ \text{death} / \text{SAE} /
-      ((\text{max pSOFA} > 1) \times (\text{if max pSOFA > 1: max
-            pSOFA else: days
-    hospitalized})) \]
-\end{itemize}
-
-Aside: \citet{wittkowskiCombiningSeveralOrdinal2004} consider combining
-ordinal variables into partial orders. This makes more sense in the
-generalized pairwise comparisons framework.
-
-\begin{figure}
-  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
-  \caption{``Angina Symptom Score'' ordinal outocome develped in
-    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
-    (1) complete tie-breaking (of units of antianginal medication by
-    episodes of angina) and (2) overriding (of angina episodes and
-      angina medication by adverse clinical events; of all other outcomes
-  by death; etc.)}
-  \label{fig:rajkumar_et_al_outcome}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
-  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
-  \caption{Ordinal outcome used in the planned PASSPORT adaptive
-    platform trial. This illustrates (1) overriding (of hospital
-    outcomes by pSOFA outcomes, etc.) but also (2) using tie-breaking
-    to augment a ``state'' outcome (status at day 28) with
-    longitudinal (maximum pSOFA while hospitalized) and time-to-event
-    (days hospitalized)
-  information}
-  \label{fig:passport_outcome}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
-  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
-  \caption{Ordinal outcome used in the planned BANDICOOT adaptive
-    platform trial \citep{walkerCodesigningNovelOrdinal2025}.
-  Illustrates \textbf{parallel composition} of component outcomes}
-  \label{fig:parallel_bandicoot}
-\end{figure}
-
-\subsubsection{Tie breaking can augment a state outcome with longitudinal
-information}
-
-Distinguish between ``state'' and ``trajectory'' variables. State
-variables summarize patient outcomes at a particular moment in time.
-Trajectory variables, including times-to-event, summarize patient
-state over a timecourse.
-
-Trajectory and state variables can be incorporated into an ordinal outcome. For
-example (from PASSPORT) ``worst pSOFA while hospitalized until day 28'' and
-``length of hospitalization'', which are longitudinal variables (summary of a
-trajectory), are combined with the day 28 state variable ``death''.
-
-Incorporating trajectory variables into an ordinal outcome may seem unnatural.
-But arguably most time-to-event analyses are in fact analyses of an
-ordinal outcome. The Cox model for time-to-event data treats time as ordinal:
-because (1) the baseline hazard is not modelled, (2) the model itself is
-invariant to monotone rescalings of the time axis, and (3) the (partial)
-likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
-marginal likelihood). The Cox model is essentially an ordinal
-stopping ratio model with complementary log-log link. (Perhaps a
-difference is censoring.)
-
-\subsubsection{Overriding can augment a continuous outcome}
-
-For example overriding a continuous outcome by death. It would be
-good to have an example of this.
-
-\subsubsection{Precision and power depend mostly on the probability of the
-largest category}
-
-Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
-efficiency (variance) of an unadjusted treatment comparison using an ordinal
-versus continuous outcome in the proportional odds model is given by $1 - \sum
-\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
-over control and experimental groups) of outcome category $c$.
-
-Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
-categories when the probabilities are all equal: $\overline p_c = 1/C$.
-
-We have simulation results and some math to show that the value of $\sum_c
-\overline p_c^3$ is \emph{essentially a function of the largest category
-probability} $\overline{p_{\max}} = \max \overline{p_c}$. See \ref{fig:p_max}
-for an illustration.
-
-\begin{figure}
-  \includegraphics[width=6in]{p_max_controls_efficiency.png}
-  \caption{Bounds on the relative efficiency of an ordinal outcome
-    compared to its continuous counterpart as a function of the maximum
-    category probability $p_\text{max}$. The upper bound (best case
-    relative efficiency for a given $p_\text{max}$) is always achieved in the
-    limit of infinitely many categories ($C = \infty$). The lower bound
-    is always achieved when the number of categories is as small as
-  possible for a given $p_\text{max}$.}
-  \label{fig:p_max}
-\end{figure}
-
-Thus simple advice is: for statistical efficiency, try to make the proportion
-of patients in the largest category small.
-
-However a converse is: if the proportion of patients in the largest category is
-small, not much is gained by tie-breaking.
-
-\subsection{Coarsening and incorporating intercurrent events for validity}
-
-\subsubsection{Coarsening to avoid directly comparing levels}
-
-In the PASSPORT outcome (Figure \ref{fig:passport_outcome}), for
-example, we include both digit
-amputation and ECMO
-any time within 28 days post-enrollment under the heading of
-``adverse clinical events''.
-
-Including these outcomes within the same level of an
-ordinal outcome does not mean assuming that the two outcomes have the
-\emph{same} utility. Rather it represents \emph{avoiding comparing}
-their utilities. (This was something clinicians asked about more than once.)
-
-\subsubsection{Coarsening for simplicity}
-
-Too many levels can make it hard to directly compare meaningful
-proportions/probabilities.
-
-(It isn't really a problem for model fitting with modern software.)
-
-Coarsening, including combining levels and replacing complete
-tie-breaks with overrides, could help with this.
-
-\subsubsection{Incorporating intercurrent events using tie-breaking
-or overriding}
-
-The intercurrent event could be (1) incorporated as a level (2) used
-for tie-breaking or overriding.
-
-This is a version of the ``composite strategy'' as described in e.g.
-\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
-more efficiency than creating a \emph{binary} composite outcome.
-
-\section{Choosing ordinal effect measures}
-
-Goal: we want to compare two or more distributions over ordinal
-outcomes with a single number. (Two distributions when comparing control and
-experimental groups; more distributions when e.g. adjusting for covariates.)
-
-In general this represents a loss of information. Let's say the
-outcome has $C$ categories. Then the distribution in each
-arm is specified by $C-1$ numbers (for the first $C-1$ probabilities,
-  or cumulative logits, or stopping ratio logits--reflecting the
-sum-to-1 constraint on probability distributions) and so the
-\emph{difference} between the distributions is also specified by
-$C-1$ numbers (for example the differences between the cumulative
-logits, or logit stopping probabilities).
-
-An effect measure thus represents a way to collapse these $C-1$
-numbers into one.
-
-It usually doesn't make sense to consider effect measures based on
-numeric coding (e.g. difference in mean levels). This is adding
-information that is not present in the basic ordinal structure.
-
-\subsection{Model-based effect measures}
-
-Model-based: common odds ratio; common stopping ratio or common continuation
-ratio.
-
-What they have in common: the effect measure is a byproduct of fitting a model
-to the data.
-
-The model typically assumes that a certain pattern holds in the outcome
-distributions across groups. If the pattern holds, then the differences in
-distributions can be described by a single number (with more than two groups,
-$G-1$ numbers) instead of $(C-1)$ numbers.
-
-\subsection{Model-free or nonparametric effect measures}
-
-These are not based on a model (assumed pattern in the differences
-between outcome distributions).
-
-\subsubsection{Effect measures based on ``independent pairwise
-comparison'' construction}
-
-Probabilistic index/probability of
-superiority, win ratio, win odds, etc. These predate win ratio and DOOR methods
-\citep[e.g.][p.~14]{agrestiAnalysisOrdinalCategorical2010}.
-
-Rather they are all be based on the ``\emph{pairs construction}'',
-where we imagine independently sampling patients from the control and
-treatment groups and considering ``wins'' ``ties'' and ``losses''
-(corresponding to concordant and discordant pairs and ties in $y$
-respectively). They can
-also be \emph{estimated from data} using pairs of patients, but this
-is not essential and they could also be estimated by fitting a
-proportional odds model, for example (though they would not be a
-simple function of a coefficient).
-
-\subsubsection{Dichotomizing or comparing with a reference group}
-
-\begin{itemize}
-  \item Reducing to a binary outcome by picking a specific cutpoint,
-    e.g. ``hospitalization or
-    worse''. This is effectively using a binary outcome, but with the
-    ordinal structure used to gain power if fitting a model.
-  \item Reducing to a binary outcome using quantiles of a reference
-    group; for example,
-    ``worse than the median outcome under standard care''
-  \item Ridit analysis
-    \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
-    smithsonReceiverOperatingCharacteristic2023, jansenRiditAnalysisReview1984}
-\end{itemize}
-
-\subsection{Comparing model-based and nonparametric effect measures}
-
-Both of these can be considered ``estimands'', insofar as we consider the
-estimand associated with the model-based effect measures to be given by the
-best-fitting model in the population.
-
-Two classes of assumption violations for model-based effect measures
-are \emph{magnitude}-heterogeneity across cutpoints (for example,
-  effect of a treatment on hospital length of stay or worse is larger
-  than its effect on
-death on the relevant scale) and
-\emph{sign}-heterogeneity
-across cutpoints (failure of stochastic dominance -- for example,
-treatment is beneficial for hospital length of stay, harmful for death).
-
-Nonparametric effect measures may appear to be assumption-free. However
-
-\begin{itemize}
-  \item Assumptions don't need to hold exactly for the model-based
-    effect measures to make sense
-  \item In particular, if stochastic dominance holds, we can think of
-    both model-based and nonparametric effect measures as
-    testing the null hypothesis of no effect ``in a particular
-    direction''. A weak interpretation of any particular
-    effect measure
-    is that it supplies this ``direction''.
-
-    For example the Wilcoxon test is
-    equivalent to the score test from fitting a proportional odds model.
-    Thus fitting a proportional odds model (or any other model) has a
-    valid null-testing
-    interpretation even if proportional odds doesn't hold.
-  \item Under failure of stochastic dominance (``sign
-    heterogeneity''), \emph{neither} parametric nor nonparametric
-    effect measures make sense, since it doesn't make sense to
-    summarize effects by a single number. Need to dig into the
-    cutpoint-specific effects.
-  \item Model-free methods don't define a common scale for comparing
-    more than two interventions. For example, model-free effect
-    measures can be \emph{nontransitive}, so it's not clear how to
-    use them in $\geq 2$-arm studies
-  \item For the same reason, model-based methods have additional
-    advantages, for example
-    the possibility of adjustment for covariates.
-
-    When additivity approximately holds, this yields a ``best of both
-    worlds'' effect measure that is
-    personalized (conditional) yet reportable as a single
-    number.
-\end{itemize}
-
-Harrell critiques \citep{harrellOverviewCompositeOutcome2024,
-harrellRareDegenerativeDiseases2024, harrellViewsCompositeOutcome}: mainly that
-win probability etc. are \emph{relative measures}:
-
-\begin{itemize}
-  \item Comparable to Cohen's $d$ or $t$-statistic (difference in
-    means divided by standard deviation or standard error) rather than
-    difference in means: compare: standardized effect size of 0.2 vs.
-    difference in means of 10 mmHg
-  \item Affected by study inclusion criteria (why? probably needs to
-    be fleshed out.)
-\end{itemize}
-
-Simulations show a typically very strong association between
-proportional odds ratios and $c$-index
-\citep{harrellViolationProportionalOdds2020}.
-
-One point of view is that one would ``ideally'' analyze ordinal
-outcomes by reporting the associated expected utilities \todo{cite Berry
-talk; example of utility-weighted Rankin scale} and that since
-existing ordinal methods supply such utilities implicitly why not
-just supply them explicitly. A counterpoint is that the methods do not, in fact,
-supply utilities implicitly.
-
-We should distinguish between effect measures used for \emph{analysis} and for
-\emph{reporting}. It may make sense to report the probabilistic index or a
-single-cutpoint binary outcome even if the analysis is undertaken using a
-proportional odds model.
-
-\section{Situating win/DOOR/etc. methods in the broader context of ordinal
-outcomes and effect measures}
-
-Terms to define/distinguish:
-\begin{itemize}
-  \item Desirability of Outcome Ranking (outcome)
-  \item Hierarchical composite endpoint (outcome)
-  \item Prioritized outcomes (outcome)
-  \item Win methods (win probability, win odds, win ratio) (effect
-    measure; estimation methods)
-  \item Generalized pairwise comparisons (effect measure)
-\end{itemize}
-
-These approaches implicitly \textbf{bundle together three different things}:
-\begin{enumerate}
-  \item Constructing an implicit ordinal outcome through
-    complete tie-breaking
-  \item Targeting a nonparametric effect measure defined via the
-    ``pairs'' device (probabilistic
-    index/win ratio/win odds/etc.)
-  \item Estimating that effect measure in a nonparametric/flexible
-    manner using pairs, often also with devices for accounting for censoring
-\end{enumerate}
-
-This paper is focused on first two steps.
-
-Why is it useful to consider these three steps separately?
-
-Consider some of the critiques of win methods in
-\citet[e.g.~][]{ajufoFallaciesUsingWin2023}
-
-\begin{itemize}
-  \item Situating methods in their proper context can help
-    trialists make informed choices among them
-  \item Win methods can make use of
-    plots and diagnostics for general ordinal outcomes, e.g.
-    comparative CDF plots.
-  \item Ordinal regression models can use win methods for model checking; it is
-    also possible to present a model-based win effect measure alongside a common
-    odds ratio for example.
-  \item The outcomes \emph{implicitly} constructed using
-    DOOR and win methods may contain many levels, and may contain
-    levels where an
-    effect is not so meaningful (e.g. time-to-death when the followup time is
-    short). By \emph{explicitly} constructing the associated ordinal outcome one
-    might be forced to be more aware of this
-  \item What is a reasonable win
-    analysis of a specific component score? From the traditional
-    ordinal endpoint
-    school of thought what would make most sense is to \emph{only
-      tie-break for the
-    lowest and highest levels of that score}, and e.g. plot eCDFs of this new
-    variable. Is this what the plots do that break down wins and losses by
-    component score?
-\end{itemize}
-
-Disadvantages of generalized pairwise comparison approaches
-\begin{itemize}
-  \item Inherited from downsides of complete tiebreaking
-  \item Inherited from downsides of nonparametric/pairwise effect measures
-  \item Regression methods exist, but are newer and less battle-tested
-\end{itemize}
-
-\todo{``Partial credit'' analysis?}
-
-\section{Conclusions}
-
-\newpage
-
-\section{Bibliography}
-
-\printbibliography
-
 \end{document}

### 2025-05-29 choosing_ordinal_effect_measures.tex
Simpler names
Begin splitting into two logical manuscripts

@@
 
-\addbibresource{bibliography.bib}
+\addbibresource{references.bib}
 \graphicspath{{fig/}}
@@
 
+\section{Desiderata for ordinal effect measures}
+
+What about effect measures? We generally seek effect measures that
+\begin{enumerate}
+  \item accurately summarize decision-relevant differences in distributions of
+    patient benefit and harm
+  \item are easy to interpret and to translate into practical recommendations
+  \item can be incorporated into statistical models where required
+  \item are easy to generalize or ``transport'' across contexts or settings
+\end{enumerate}
+
+Key references:
+\citet{agrestiMeasuresNominalOrdinalAssociation1981,agrestiOrdinalProbabilityEffect2017a,
+agrestiSimpleWaysInterpret2018a}
+
+Goal: we want to compare two or more distributions over ordinal
+outcomes with a single number. (Two distributions when comparing control and
+experimental groups; more distributions when e.g. adjusting for covariates.)
+
+In general this represents a loss of information. Let's say the
+outcome has $C$ categories. Then the distribution in each
+arm is specified by $C-1$ numbers (for the first $C-1$ probabilities,
+  or cumulative logits, or stopping ratio logits--reflecting the
+sum-to-1 constraint on probability distributions) and so the
+\emph{difference} between the distributions is also specified by
+$C-1$ numbers (for example the differences between the cumulative
+logits, or logit stopping probabilities).
+
+An effect measure thus represents a way to collapse these $C-1$
+numbers into one.
+
+It usually doesn't make sense to consider effect measures based on
+numeric coding (e.g. difference in mean levels). This is adding
+information that is not present in the basic ordinal structure.
+
+\section{A taxonomy of ordinal effect measures}
+
+\subsection{Model-based effect measures}
+
+\begin{figure}
+  \includegraphics[width=7in]{illustrating_common_or.pdf}
+  \caption{The common odds ratio, a model-based effect measure, comes
+    from the best-fitting parallel curves to the logit transformed
+    empirical cumulative distribution. \textbf{Ideally would illustrate
+      using toy bar charts as above, with only 2 groups? Although it's a
+  strength that model-based measures can compare multiple groups.}}
+  \label{fig:illustrating_or}
+\end{figure}
+
+\begin{itemize}
+  \item
+    common odds ratio (or e.g. probit analogue; there are versions of
+    this for any cumulative link)
+  \item common stopping ratio or common continuation
+    ratio (hazard ratio is an example with implicitly ordinal time)
+\end{itemize}
+
+What they have in common: the effect measure is a byproduct of fitting a model
+to the data. See Figure \ref{fig:illustrating_or} for an illustration
+for the common odds ratio.
+
+A good overview of ordinal models is \citet{burknerOrdinalRegressionModels2019}.
+
+The model assumes that a certain \emph{pattern} is present in the outcome
+distributions across groups. If the pattern ``holds'', then the differences in
+distributions can be described by a single number (with more than two groups,
+$G-1$ numbers) instead of $(C-1)$ numbers.
+
+In Figure \ref{fig:illustrating_or} the pattern assumed to hold is
+that the logit-transformed eCDFs in each group are parallel.
+
+\subsection{Model-free or nonparametric effect measures}
+
+These are not based on a model (assumed pattern in the differences
+between outcome distributions).
+
+\subsubsection{Effect measures based on ``independent pairs'' or
+``win'' constructions}
+
+\begin{figure}
+  \includegraphics[width=7.5in]{effect_measures_pp_plot.pdf}
+  \caption{Illustrating independent pairwise comparison-based effect
+    measures using a PP-plot or ROC curve. Adapted from
+  \citet{smithsonReceiverOperatingCharacteristic2023}.}
+  \label{fig:illustrating_pairwise_comparisons}
+\end{figure}
+
+Probabilistic index/probability of
+superiority, win ratio, win odds, etc. These predate win ratio and
+DOOR methods
+\citep[e.g.][p.~14]{agrestiAnalysisOrdinalCategorical2010}.
+
+Rather they are all be based on the ``\emph{pairs construction}'',
+where we imagine independently sampling patients from the control and
+treatment groups and considering ``wins'' ``ties'' and ``losses''
+(corresponding to concordant and discordant pairs and ties in $y$
+respectively). See Figure \ref{fig:illustrating_pairwise_comparisons}
+for an illustration.
+
+They can
+also be \emph{estimated from data} using pairs of patients, but this
+is not essential and they could also be estimated by fitting a
+proportional odds model, for example (though they would not be a
+simple function of a coefficient).
+
+\subsubsection{Dichotomizing or comparing with a reference group}
+
+\begin{itemize}
+  \item Dichotomizing by picking a specific cutpoint,
+    e.g. ``hospitalization or
+    worse''. This is effectively using a binary outcome, but with the
+    ordinal structure used to gain power if fitting a model.
+  \item Dichotomizing using quantiles of a reference
+    group; for example,
+    ``worse than the median outcome under standard care''
+  \item Simultaneously comparing all quantiles to those of a
+    reference group: ridit analysis
+    \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
+      smithsonReceiverOperatingCharacteristic2023,
+    jansenRiditAnalysisReview1984}
+\end{itemize}
+
+\subsection{Comparing model-based and nonparametric effect measures}
+
+Both of these can be considered ``estimands'', insofar as we consider the
+estimand associated with the model-based effect measures to be given by the
+best-fitting model in the population.
+
+Two classes of assumption violations for model-based effect measures
+are \emph{magnitude}-heterogeneity across cutpoints (for example,
+  effect of a treatment on hospital length of stay or worse is larger
+  than its effect on
+death on the relevant scale) and
+\emph{sign}-heterogeneity
+across cutpoints (failure of stochastic dominance -- for example,
+treatment is beneficial for hospital length of stay, harmful for death).
+
+Nonparametric effect measures may appear to be assumption-free. However
+
+\begin{itemize}
+  \item Assumptions don't need to hold exactly for the model-based
+    effect measures to make sense
+  \item In particular, if stochastic dominance holds, we can think of
+    both model-based and nonparametric effect measures as
+    testing the null hypothesis of no effect ``in a particular
+    direction''. A weak interpretation of any particular
+    effect measure
+    is that it supplies this ``direction''.
+
+    For example the Wilcoxon test is
+    equivalent to the score test from fitting a proportional odds model.
+    Thus fitting a proportional odds model (or any other model) has a
+    valid null-testing
+    interpretation even if proportional odds doesn't hold.
+  \item Under failure of stochastic dominance (``sign
+    heterogeneity''), \emph{neither} parametric nor nonparametric
+    effect measures make sense, since it doesn't make sense to
+    summarize effects by a single number. Need to dig into the
+    cutpoint-specific effects.
+  \item Model-free methods don't define a common scale for comparing
+    more than two interventions. For example, model-free effect
+    measures can be \emph{nontransitive}, so it's not clear how to
+    use them in $\geq 2$-arm studies
+  \item For the same reason, model-based methods have additional
+    advantages, for example
+    the possibility of adjustment for covariates.
+
+    When additivity approximately holds, this yields a ``best of both
+    worlds'' effect measure that is
+    personalized (conditional) yet reportable as a single
+    number.
+\end{itemize}
+
+Harrell critiques \citep{harrellOverviewCompositeOutcome2024,
+harrellRareDegenerativeDiseases2024, harrellViewsCompositeOutcome}:
+mainly that
+win probability etc. are \emph{relative measures}:
+
+\begin{itemize}
+  \item Comparable to Cohen's $d$ or $t$-statistic (difference in
+    means divided by standard deviation or standard error) rather than
+    difference in means: compare: standardized effect size of 0.2 vs.
+    difference in means of 10 mmHg
+  \item Affected by study inclusion criteria (why? probably needs to
+    be fleshed out.)
+\end{itemize}
+
+Simulations show a typically very strong association between
+proportional odds ratios and $c$-index
+\citep{harrellViolationProportionalOdds2020}.
+
+We should distinguish between effect measures used for \emph{analysis} and for
+\emph{reporting}. It may make sense to report the probabilistic index or a
+single-cutpoint binary outcome even if the analysis is undertaken using a
+proportional odds model.
+
+\newpage
+
+\printbibliography
+
 \end{document}

### 2025-05-29 constructing_ordinal_outcomes.tex
Rename + start effect measures draft
Simpler names
Begin splitting into two logical manuscripts

--- 2025-05-15 paper.tex
@@
 
-\addbibresource{bibliography.bib}
+\addbibresource{references.bib}
 \graphicspath{{fig/}}
 
-\title{Guidance for clinicians in constructing ordinal outcomes and choosing
-ordinal effect measures for clinical trials}
+\title{Constructing ordinal outcomes for clinical trials}
 \author{Leon Di Stefano + Ordinal
@@
 
-\section{Desiderata for outcomes and effect measures: validity and efficiency}
+\subsection{Desiderata for (ordinal) outcomes: validity and efficiency}
 
@@
 
-An example of an outcome which is precise but which may be invalid
+An example of an outcome which is efficient but which may be invalid
 might be continuously-measured concentrations of a biomarker which is a poor
@@
 
-What about effect measures? We generally seek effect measures that
-\begin{enumerate}
-  \item accurately summarize decision-relevant differences in distributions of
-    patient benefit and harm
-  \item are easy to interpret and to translate into practical recommendations
-  \item can be incorporated into statistical models where required
-  \item are easy to generalize or ``transport'' across contexts or settings
-\end{enumerate}
+One point of view is that one would ``ideally'' analyze ordinal outcomes by
+reporting the associated expected utilities \todo{cite Berry talk; example of
+utility-weighted Rankin scale} and that since existing ordinal methods supply
+such utilities implicitly why not just supply them explicitly. A counterpoint
+is that the methods do not, in fact, supply utilities implicitly.
+Another is that ordinal outcomes let one ``hedge'' about precise
+utillities and this is OK.
 
-\section{Combining and coarsening ordinal outcomes}
+\subsection{Constructing versus validating}
 
-\subsection{3 ways to combine/fine-grain ordinal outcomes: sequential,
+\section{Ways to combine ordinal outcomes}
+
+\subsection{Sequential,
 hierarchical, \& parallel composition}
@@
 
-\subsubsection{Another way: scores}
-
-Treat each input factor as
-numeric then sum them up. I believe \citet{ongUnlockingDOORHow2023}
-has examples?
-
 \begin{figure}
@@
     platform trial \citep{walkerCodesigningNovelOrdinal2025}.
-  Illustrates \textbf{parallel composition} of component outcomes}
+  Illustrates \textbf{parallel composition} of component outcomes.}
   \label{fig:parallel_bandicoot}
@@
 
-\subsection{Reasons to combine ordinal outcomes}
+\subsection{Converting back to numeric scores}
 
-\subsubsection{Incorporating intercurrent events for validity}
+Treat each input factor as
+numeric then sum them up. I believe \citet{ongUnlockingDOORHow2023}
+has examples?
 
-The intercurrent event could be (1) incorporated as a level (2) used
-for tie-breaking or overriding.
+This doesn't respect the ordinal nature of the variables. That is, it
+implicitly uses information about the \emph{differences} between levels.
 
-This is a version of the ``composite strategy'' as described in e.g.
-\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
-more efficiency than creating a \emph{binary} composite outcome.
-
-For example overriding a numeric outcome by death.
-Example: REMAP-CAP: organ-free survival, overridden by death
-(Figure \ref{fig:remap_cap_override}).
-
-\begin{figure}
-  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
-  \caption{Ordinal outcome used in the REMAP-CAP adaptive
-    platform trial
-    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
-    Illustrates overriding a numeric ordinal outcome (organ
-    support-free days) by a binary ordinal outcome (death; this is
-    ``sequential composition''), which
-  would also be an intercurrent event.}
-  \label{fig:remap_cap_override}
-\end{figure}
-
-\subsubsection{Increasing precision by reducing the probability of the
-largest category}
-
-Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
-efficiency (variance) of an unadjusted treatment comparison using an ordinal
-versus continuous outcome in the proportional odds model is given by $1 - \sum
-\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
-over control and experimental groups) of outcome category $c$.
-
-Same should apply to the Mann-Whitney-Wilcoxon test, and therefore to
-the win probability, and therefore at least approximately to win
-difference/win ratio/win odds.
-
-Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
-categories when the probabilities are all equal: $\overline p_c = 1/C$.
-
-We have simulation results and some math to show that the value of $\sum_c
-\overline p_c^3$ is \emph{essentially a function of the largest category
-probability} $\overline{p_{\max}} = \max \overline{p_c}$. See Figure
-\ref{fig:p_max} for an illustration.
-
-\begin{figure}
-  \includegraphics[width=6in]{p_max_controls_efficiency.png}
-  \caption{Bounds on the relative efficiency of an ordinal outcome
-    compared to its continuous counterpart as a function of the maximum
-    category probability $p_\text{max}$. The upper bound (best case
-    relative efficiency for a given $p_\text{max}$) is always achieved in the
-    limit of infinitely many categories ($C = \infty$). The lower bound
-    is always achieved when the number of categories is as small as
-  possible for a given $p_\text{max}$.}
-  \label{fig:p_max}
-\end{figure}
-
-Thus simple advice is: for statistical efficiency, try to make the proportion
-of patients in the largest category small.
-
-However a converse is: if the proportion of patients in the largest
-category is
-small, not much is gained by tie-breaking.
-
-\subsubsection{Augmenting a state outcome with longitudinal
-information}
-
-Distinguish between ``state'' and ``trajectory'' variables. State
-variables summarize patient outcomes at a particular moment in time.
-Trajectory variables, including times-to-event, summarize patient
-state over a timecourse.
-
-Trajectory and state variables can be incorporated into an ordinal outcome. For
-example (from PASSPORT) ``worst pSOFA while hospitalized until day 28'' and
-``length of hospitalization'', which are longitudinal variables (summary of a
-trajectory), are combined with the day 28 state variable ``death''.
-
-Incorporating trajectory variables into an ordinal outcome may seem unnatural.
-But arguably most time-to-event analyses are in fact analyses of an
-ordinal outcome. The Cox model for time-to-event data treats time as ordinal:
-because (1) the baseline hazard is not modelled, (2) the model itself is
-invariant to monotone rescalings of the time axis, and (3) the (partial)
-likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
-marginal likelihood). The Cox model is essentially an ordinal
-stopping ratio model with complementary log-log link. (Perhaps a
-difference is censoring.)
-
-\subsection{Reasons to coarsen ordinal outcomes}
-
-\subsubsection{Coarsening to avoid directly comparing levels}
-
-In the PASSPORT outcome (Figure \ref{fig:passport_outcome}), for
-example, we include both digit
-amputation and ECMO
-any time within 28 days post-enrollment under the heading of
-``adverse clinical events''.
-
-Including these outcomes within the same level of an
-ordinal outcome does not mean assuming that the two outcomes have the
-\emph{same} utility. Rather it represents \emph{avoiding comparing}
-their utilities. (This was something clinicians asked about more than once.)
-
-Also, we "chunk" longer hospital stays more aggressively to avoid
-small categories, for interpretability's sake.
-
-\subsubsection{Coarsening for ease of communication}
-
-Too many levels can make it hard to directly compare meaningful
-proportions/probabilities.
-
-(It isn't really a problem for model fitting with modern software.)
-
-Coarsening, including combining levels and replacing complete
-tie-breaks with overrides, could help with this.
-
-\section{Choosing ordinal effect measures}
-
-Key references:
-\citet{agrestiMeasuresNominalOrdinalAssociation1981,agrestiOrdinalProbabilityEffect2017a,
-agrestiSimpleWaysInterpret2018a}
-
-Goal: we want to compare two or more distributions over ordinal
-outcomes with a single number. (Two distributions when comparing control and
-experimental groups; more distributions when e.g. adjusting for covariates.)
-
-In general this represents a loss of information. Let's say the
-outcome has $C$ categories. Then the distribution in each
-arm is specified by $C-1$ numbers (for the first $C-1$ probabilities,
-  or cumulative logits, or stopping ratio logits--reflecting the
-sum-to-1 constraint on probability distributions) and so the
-\emph{difference} between the distributions is also specified by
-$C-1$ numbers (for example the differences between the cumulative
-logits, or logit stopping probabilities).
-
-An effect measure thus represents a way to collapse these $C-1$
-numbers into one.
-
-It usually doesn't make sense to consider effect measures based on
-numeric coding (e.g. difference in mean levels). This is adding
-information that is not present in the basic ordinal structure.
-
-\subsection{Model-based effect measures}
-
-\begin{figure}
-  \includegraphics[width=7in]{illustrating_common_or.pdf}
-  \caption{The common odds ratio, a model-based effect measure, comes
-    from the best-fitting parallel curves to the logit transformed
-    empirical cumulative distribution. \textbf{Ideally would illustrate
-      using toy bar charts as above, with only 2 groups? Although it's a
-  strength that model-based measures can compare multiple groups.}}
-  \label{fig:illustrating_or}
-\end{figure}
-
-Model-based: common odds ratio; common stopping ratio or common continuation
-ratio.
-
-What they have in common: the effect measure is a byproduct of fitting a model
-to the data. See Figure \ref{fig:illustrating_or} for an illustration
-for the common odds ratio.
-
-The model typically assumes that a certain pattern holds in the outcome
-distributions across groups. If the pattern holds, then the differences in
-distributions can be described by a single number (with more than two groups,
-$G-1$ numbers) instead of $(C-1)$ numbers.
-
-In Figure \ref{fig:illustrating_or} the pattern assumed to hold is
-that the logit-transformed eCDFs in each group are parallel.
-
-\subsection{Model-free or nonparametric effect measures}
-
-These are not based on a model (assumed pattern in the differences
-between outcome distributions).
-
-\subsubsection{Effect measures based on ``independent pairwise
-comparison'' construction}
-
-\begin{figure}
-  \includegraphics[width=7.5in]{effect_measures_pp_plot.pdf}
-  \caption{Illustrating independent pairwise comparison-based effect
-    measures using a PP-plot or ROC curve. Adapted from
-  \citet{smithsonReceiverOperatingCharacteristic2023}.}
-  \label{fig:illustrating_pairwise_comparisons}
-\end{figure}
-
-Probabilistic index/probability of
-superiority, win ratio, win odds, etc. These predate win ratio and
-DOOR methods
-\citep[e.g.][p.~14]{agrestiAnalysisOrdinalCategorical2010}.
-
-Rather they are all be based on the ``\emph{pairs construction}'',
-where we imagine independently sampling patients from the control and
-treatment groups and considering ``wins'' ``ties'' and ``losses''
-(corresponding to concordant and discordant pairs and ties in $y$
-respectively). See Figure \ref{fig:illustrating_pairwise_comparisons}
-for an illustration.
-
-They can
-also be \emph{estimated from data} using pairs of patients, but this
-is not essential and they could also be estimated by fitting a
-proportional odds model, for example (though they would not be a
-simple function of a coefficient).
-
-\subsubsection{Dichotomizing or comparing with a reference group}
-
-\begin{itemize}
-  \item Dichotomizing by picking a specific cutpoint,
-    e.g. ``hospitalization or
-    worse''. This is effectively using a binary outcome, but with the
-    ordinal structure used to gain power if fitting a model.
-  \item Dichotomizing using quantiles of a reference
-    group; for example,
-    ``worse than the median outcome under standard care''
-  \item Simultaneously comparing all quantiles to those of a
-    reference group: ridit analysis
-    \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
-      smithsonReceiverOperatingCharacteristic2023,
-    jansenRiditAnalysisReview1984}
-\end{itemize}
-
-\subsection{Comparing model-based and nonparametric effect measures}
-
-Both of these can be considered ``estimands'', insofar as we consider the
-estimand associated with the model-based effect measures to be given by the
-best-fitting model in the population.
-
-Two classes of assumption violations for model-based effect measures
-are \emph{magnitude}-heterogeneity across cutpoints (for example,
-  effect of a treatment on hospital length of stay or worse is larger
-  than its effect on
-death on the relevant scale) and
-\emph{sign}-heterogeneity
-across cutpoints (failure of stochastic dominance -- for example,
-treatment is beneficial for hospital length of stay, harmful for death).
-
-Nonparametric effect measures may appear to be assumption-free. However
-
-\begin{itemize}
-  \item Assumptions don't need to hold exactly for the model-based
-    effect measures to make sense
-  \item In particular, if stochastic dominance holds, we can think of
-    both model-based and nonparametric effect measures as
-    testing the null hypothesis of no effect ``in a particular
-    direction''. A weak interpretation of any particular
-    effect measure
-    is that it supplies this ``direction''.
-
-    For example the Wilcoxon test is
-    equivalent to the score test from fitting a proportional odds model.
-    Thus fitting a proportional odds model (or any other model) has a
-    valid null-testing
-    interpretation even if proportional odds doesn't hold.
-  \item Under failure of stochastic dominance (``sign
-    heterogeneity''), \emph{neither} parametric nor nonparametric
-    effect measures make sense, since it doesn't make sense to
-    summarize effects by a single number. Need to dig into the
-    cutpoint-specific effects.
-  \item Model-free methods don't define a common scale for comparing
-    more than two interventions. For example, model-free effect
-    measures can be \emph{nontransitive}, so it's not clear how to
-    use them in $\geq 2$-arm studies
-  \item For the same reason, model-based methods have additional
-    advantages, for example
-    the possibility of adjustment for covariates.
-
-    When additivity approximately holds, this yields a ``best of both
-    worlds'' effect measure that is
-    personalized (conditional) yet reportable as a single
-    number.
-\end{itemize}
-
-Harrell critiques \citep{harrellOverviewCompositeOutcome2024,
-harrellRareDegenerativeDiseases2024, harrellViewsCompositeOutcome}:
-mainly that
-win probability etc. are \emph{relative measures}:
-
-\begin{itemize}
-  \item Comparable to Cohen's $d$ or $t$-statistic (difference in
-    means divided by standard deviation or standard error) rather than
-    difference in means: compare: standardized effect size of 0.2 vs.
-    difference in means of 10 mmHg
-  \item Affected by study inclusion criteria (why? probably needs to
-    be fleshed out.)
-\end{itemize}
-
-Simulations show a typically very strong association between
-proportional odds ratios and $c$-index
-\citep{harrellViolationProportionalOdds2020}.
-
-One point of view is that one would ``ideally'' analyze ordinal
-outcomes by reporting the associated expected utilities \todo{cite Berry
-talk; example of utility-weighted Rankin scale} and that since
-existing ordinal methods supply such utilities implicitly why not
-just supply them explicitly. A counterpoint is that the methods do
-not, in fact,
-supply utilities implicitly.
-
-We should distinguish between effect measures used for \emph{analysis} and for
-\emph{reporting}. It may make sense to report the probabilistic index or a
-single-cutpoint binary outcome even if the analysis is undertaken using a
-proportional odds model.
-
-\section{Situating win/DOOR/etc. methods in the broader context of ordinal
-outcomes and effect measures}
+\subsection{Pairwise comparison methods (win, DOOR) implicitly construct ordinal
+outcomes}
 
@@
 
-This paper is focused on first two steps.
+This paper is focused on the first of these steps only.
 
-Why is it useful to consider these three steps separately?
+Why is it useful to consider the three steps separately?
 
@@
 
-\todo{``Partial credit'' analysis?}
+\section{Reasons to combine ordinal outcomes}
+
+\subsection{Providing a more holistic measure of benefit and harm}
+
+\subsection{Incorporating intercurrent events, especially death,
+for validity}
+
+The intercurrent event could be (1) incorporated as a level (2) used
+for tie-breaking or overriding.
+
+This is a version of the ``composite strategy'' as described in e.g.
+\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
+more efficiency than creating a \emph{binary} composite outcome.
+
+For example overriding a numeric outcome by death.
+Example: REMAP-CAP: organ-free survival, overridden by death
+(Figure \ref{fig:remap_cap_override}).
+
+\begin{figure}
+  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
+  \caption{Ordinal outcome used in the REMAP-CAP adaptive
+    platform trial
+    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
+    Illustrates overriding a numeric ordinal outcome (organ
+    support-free days) by a binary ordinal outcome (death; this is
+    ``sequential composition''), which
+    would also be an intercurrent event. This is another example of
+  sequential composition.}
+  \label{fig:remap_cap_override}
+\end{figure}
+
+\subsection{Increasing precision by reducing the probability of the
+largest category}
+
+Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
+efficiency (variance) of an unadjusted treatment comparison using an ordinal
+versus continuous outcome in the proportional odds model is given by $1 - \sum
+\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
+over control and experimental groups) of outcome category $c$.
+
+Same should apply to the Mann-Whitney-Wilcoxon test, and therefore to
+the win probability, and therefore at least approximately to win
+difference/win ratio/win odds.
+
+Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
+categories when the probabilities are all equal: $\overline p_c = 1/C$.
+
+We have simulation results and some maths to show that the value of $\sum_c
+\overline p_c^3$ is \emph{essentially a function of the largest category
+probability} $\overline{p_{\max}} = \max \overline{p_c}$. See Figure
+\ref{fig:p_max} for an illustration of the maths.
+
+\begin{figure}
+  \includegraphics[width=6in]{p_max_controls_efficiency.png}
+  \caption{Bounds on the relative efficiency of an ordinal outcome
+    compared to its continuous counterpart as a function of the maximum
+    category probability $p_\text{max}$. The upper bound (best case
+    relative efficiency for a given $p_\text{max}$) is always achieved in the
+    limit of infinitely many categories ($C = \infty$). The lower bound
+    is always achieved when the number of categories is as small as
+  possible for a given $p_\text{max}$.}
+  \label{fig:p_max}
+\end{figure}
+
+Thus simple advice is: for statistical efficiency, try to make the proportion
+of patients in the largest category small.
+
+However a converse is: if the proportion of patients in the largest
+category is
+small, not much is gained by tie-breaking.
+
+\subsection{Augmenting a state outcome with longitudinal
+information}
+
+Distinguish between ``state'' and ``trajectory'' variables. State
+variables summarize patient outcomes at a particular moment in time.
+Trajectory variables, including times-to-event, summarize patient
+state over a timecourse.
+
+Trajectory and state variables can be incorporated into an ordinal outcome. For
+example (from PASSPORT) ``worst pSOFA while hospitalized until day 28'' and
+``length of hospitalization'', which are longitudinal variables (summary of a
+trajectory), are combined with the day 28 state variable ``death''.
+
+Incorporating trajectory variables into an ordinal outcome may seem unnatural.
+But arguably most time-to-event analyses are in fact analyses of an
+ordinal outcome. The Cox model for time-to-event data treats time as ordinal:
+because (1) the baseline hazard is not modelled, (2) the model itself is
+invariant to monotone rescalings of the time axis, and (3) the (partial)
+likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
+marginal likelihood). The Cox model is essentially an ordinal
+stopping ratio model with complementary log-log link. (Perhaps a
+difference is censoring.)
+
+\section{Reasons to coarsen ordinal outcomes}
+
+\subsection{Coarsening to avoid directly comparing levels}
+
+In the PASSPORT outcome (Figure \ref{fig:passport_outcome}), for
+example, we include both digit
+amputation and ECMO
+any time within 28 days post-enrollment under the heading of
+``adverse clinical events''.
+
+Including these outcomes within the same level of an
+ordinal outcome does not mean assuming that the two outcomes have the
+\emph{same} utility. Rather it represents \emph{avoiding comparing}
+their utilities. (This was something clinicians asked about more than once.)
+
+Also, we "chunk" longer hospital stays more aggressively to avoid
+small categories, for interpretability's sake.
+
+\subsection{Coarsening for ease of communication}
+
+Too many levels can make it hard to directly compare meaningful
+proportions/probabilities.
+
+(It isn't really a problem for model fitting with modern software.)
+
+Coarsening, including combining levels and replacing complete
+tie-breaks with overrides, could help with this.
 
@@
 
-\section{Bibliography}
-
 \printbibliography

### 2025-06-06 choosing_ordinal_effect_measures.tex
Bunch of work including joint PO figures

@@
 
-What about effect measures? We generally seek effect measures that
+Effect measures compare distributions of outcomes across groups
+under different interventions or exposures.
+
+We generally seek effect measures that
 \begin{enumerate}
-  \item accurately summarize decision-relevant differences in distributions of
-    patient benefit and harm
+  \item correspond most directly to patient benefit/harm
   \item are easy to interpret and to translate into practical recommendations
@@
 
-Key references:
-\citet{agrestiMeasuresNominalOrdinalAssociation1981,agrestiOrdinalProbabilityEffect2017a,
-agrestiSimpleWaysInterpret2018a}
+Key references for ordinal effect measures:
+\citet{agrestiMeasuresNominalOrdinalAssociation1981,agrestiOrdinalProbabilityEffect2017,
+agrestiSimpleWaysInterpret2018}.
 
-Goal: we want to compare two or more distributions over ordinal
-outcomes with a single number. (Two distributions when comparing control and
-experimental groups; more distributions when e.g. adjusting for covariates.)
+We want to compare two or more distributions over ordinal
+outcomes with a single number. We want to compare \emph{two} distributions
+when when there are two interventions (call them control and
+experimental). But in multi-arm studies or when e.g. adjusting for
+covariates, it may become important to compare more than 2 groups
+(see the transitivity in Section \ref{sec:comparing}).
 
-In general this represents a loss of information. Let's say the
+Summarizing the comparison between two distributions with a single
+number represents a loss of information. Let's say the
 outcome has $C$ categories. Then the distribution in each
-arm is specified by $C-1$ numbers (for the first $C-1$ probabilities,
+arm is non-redundantly specified by $C-1$ numbers (for the first
+  $C-1$ probabilities,
   or cumulative logits, or stopping ratio logits--reflecting the
 sum-to-1 constraint on probability distributions) and so the
-\emph{difference} between the distributions is also specified by
+\emph{difference} between the distributions is also non-redundantly specified by
 $C-1$ numbers (for example the differences between the cumulative
-logits, or logit stopping probabilities).
+  logits, or logit stopping probabilities -- these latter numbers are
+unconstrained).
 
-An effect measure thus represents a way to collapse these $C-1$
-numbers into one.
+Since an effect measure thus represents a way to collapse these $C-1$
+numbers into one, it must lose information.
 
-It usually doesn't make sense to consider effect measures based on
-numeric coding (e.g. difference in mean levels). This is adding
-information that is not present in the basic ordinal structure.
+Truly ordinal effect measures depend only on the ordinal structure of
+the outcomes, rather than, for example, a particular numeric coding
+(e.g. difference in mean levels). This is adding
+information that is not present in the ordinal structure.
 
-\section{A taxonomy of ordinal effect measures}
-
-\subsection{Model-based effect measures}
+\section{Model-based effect measures}
 
@@
     this for any cumulative link)
-  \item common stopping ratio or common continuation
-    ratio (hazard ratio is an example with implicitly ordinal time)
+  \item Others: common stopping ratio or common continuation
+    ratio (hazard ratio is an example with implicitly ordinal time);
+    see
+    \citet{burknerOrdinalRegressionModels2019,agrestiAnalysisOrdinalCategorical2010}
+    for more.
 \end{itemize}
@@
 What they have in common: the effect measure is a byproduct of fitting a model
-to the data. See Figure \ref{fig:illustrating_or} for an illustration
+to the data.
+The model assumes that a certain \emph{pattern} is present in the outcome
+distributions across groups. If the pattern (approximately) holds,
+then the differences in
+distributions can (approximately) be described by a single number.
+See Figure \ref{fig:illustrating_or} for an illustration
 for the common odds ratio.
 
-A good overview of ordinal models is \citet{burknerOrdinalRegressionModels2019}.
-
-The model assumes that a certain \emph{pattern} is present in the outcome
-distributions across groups. If the pattern ``holds'', then the differences in
-distributions can be described by a single number (with more than two groups,
-$G-1$ numbers) instead of $(C-1)$ numbers.
-
-In Figure \ref{fig:illustrating_or} the pattern assumed to hold is
+Usually model-based effect measures can put multiple groups on a
+common ``scale''. This is important when adjusting for prognostic
+factors, for example.
+A proportional odds model does this with the scale defined by a
+common set of cutpoints
+or ``intercepts''.
+Figure \ref{fig:illustrating_or} illustrates this; the pattern
+assumed to hold is
 that the logit-transformed eCDFs in each group are parallel.
 
-\subsection{Model-free or nonparametric effect measures}
+\section{Model-free or nonparametric effect measures}
 
@@
 
-\subsubsection{Effect measures based on ``independent pairs'' or
+\subsection{Effect measures based on ``independent pairs'' or
 ``win'' constructions}
@@
 \begin{figure}
-  \includegraphics[width=7.5in]{effect_measures_pp_plot.pdf}
-  \caption{Illustrating independent pairwise comparison-based effect
-    measures using a PP-plot or ROC curve. Adapted from
+  \includegraphics[width=7in]{effect_measures_pp_plot.pdf}
+  \caption{\textbf{Illustrating independent pairwise comparison-based effect
+    measures using a PP-plot or ROC curve.}
+    The $x$ axis shows the
+    cumulative distribution function (CDF) of $y$ under the control
+    intervention; the $y$ axis shows the CDF under the experimental
+    intervention. The grey
+    curve (\emph{P-P} curve) shows how the CDFs vary with one
+    another; a curve below the diagonal indicates that the CDF of $y$
+    under the experimental treatment is lower than the CDF of $y$
+    under control, and therefore that $y$ is stochastically greater
+    under the experimental treatment. Areas in the plot correspond to
+    probabilities under
+    the ``independent pairwise comparisons'' construction: the orange
+    area corresponds to the win probability (or probabilistic index,
+    probability of superiority, common language effect size) in favor
+    of the experimental treatment. The ratio of orange and blue areas
+    is the win ratio, and if the grey area -- corresponding to tied
+    pairs under independent draws -- is allocated equally to orange
+    and blue, the ratio of areas corresponds to the win odds.
+    Adapted from
   \citet{smithsonReceiverOperatingCharacteristic2023}.}
@@
 
-Probabilistic index/probability of
-superiority, win ratio, win odds, etc. These predate win ratio and
-DOOR methods
+Probabilistic index, probability of
+superiority, common language effect size, win ratio, win odds, etc.
+These predate win ratio and
+DOOR methods and have been reinvented many times
 \citep[e.g.][p.~14]{agrestiAnalysisOrdinalCategorical2010}.
 
-Rather they are all be based on the ``\emph{pairs construction}'',
+They are all be based on the ``\emph{pairs construction}'',
 where we imagine independently sampling patients from the control and
@@
 
-\subsubsection{Dichotomizing or comparing with a reference group}
+They generalize binary effect measures and can themselves be
+generalized beyond comparing two groups to the situation where $x$ is
+also ordinal.
+
+\subsection{Effects based on the pairs constructions do not measure
+individual-level effects}
+
+Pair-construction or win methods have a subtle issue of interpretation.
+
+The win probability is \emph{not} the probability that a given
+individual's outcome under the experimental treatment is better than
+what would be obtained under the control treatment.
+
+Rather is the probability that an individual drawn at random
+from the experimental group has a better outcome than a \emph{different}
+individual drawn at random from the control group.
+
+The issue is similar to that for \emph{p}-values, which is subtly
+different from -- and often confused for -- ``the thing we want''
+(namely the posterior probability of the null hypothesis).
+
+Relatedly, I believe that \emph{win probabilities and win
+differences, unlike risks and risk differences, are non-collapsible}
+in the sense that the average of win probabilities (or differences)
+across strata is not necessarily equal to the win probability (or
+difference) in the unstratified population.
+
+\begin{figure}
+  \includegraphics[width=5in]{po_association_discrete.png}
+  \caption{\textbf{Effect measures based on the ``independent pairs''
+    construction do not refer to individual-level benefit/harm.}
+    Left: two ordinal outcome distributions. Right: different
+    possible joint distributions of potential outcomes with marginals
+    corresponding to the distributions on the left.
+    The ``probability of benefit'' -- probability that a randomly
+    selected \emph{individual} does better under the experimental
+    intervention than the control intervention -- depends on the
+    joint distribution and is higher when potential outcomes are
+    positively associated. The win probability is the probability of
+    benefit when the potential outcomes are independent (or the
+      probability that experimental intervention ``beats'' control in a
+    random \emph{pair} of individuals).
+  }
+  \label{fig:pairwise_and_pos_discrete}
+\end{figure}
+\begin{figure}
+  \includegraphics[width=7in]{po_association_continuous.png}
+  \caption{\textbf{Effect measures based on the ``independent pairs''
+    construction do not refer to individual-level benefit/harm.}
+    This figure is meant to illustrate the same point as Figure
+    \ref{fig:pairwise_and_pos_discrete}, but with
+    continuous outcomes. They make things easier because it's
+    possible to plot the
+    \emph{distribution} of individual-level treatment effects. (By
+      contrast, with ordinal outcomes there's no natural notion of the
+      difference between two levels of the scale and hence no compact way to
+    represent individual effects.) The orange criss-crosses are
+    cutpoint used to generate the data in Figure
+    \ref{fig:pairwise_and_pos_discrete}.
+    \todo{Ways to make the two plots more parallel: have green and
+      red (or blue and red) in the discrete heatmaps too; include an
+    independence case for the continuous outcomes.}
+  }
+  \label{fig:pairwise_and_pos_continuous}
+\end{figure}
+
+\subsection{Dichotomizing or using more than a single number}
+
+Ways to dichotomize:
 
@@
     group; for example,
-    ``worse than the median outcome under standard care''
-  \item Simultaneously comparing all quantiles to those of a
-    reference group: ridit analysis
+    ``proportion in the treated group with worse outcomes than the
+    median outcome under standard care''
+\end{itemize}
+
+Ways to use more than a single number to summarize effects:
+
+\begin{itemize}
+  \item Plotting cumulative distribution functions or survival functions
+  \item Plotting ridits, which compare ordinal distributions using a
+    reference group and are closely connected to win probability
     \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
@@
 
-\subsection{Comparing model-based and nonparametric effect measures}
+\section{Comparing model-based and nonparametric effect measures}
+\label{sec:comparing}
 
@@
     effect measures to make sense
+    \begin{itemize}
       \item In particular, if stochastic dominance holds, we can think of
         both model-based and nonparametric effect measures as
-    testing the null hypothesis of no effect ``in a particular
+        testing the null hypothesis of no effect ``in a specific
         direction''. A weak interpretation of any particular
@@
         For example the Wilcoxon test is
-    equivalent to the score test from fitting a proportional odds model.
+        equivalent to the score test from fitting a proportional odds model;
+        the implied ``directions from the null'' are the same.
         Thus fitting a proportional odds model (or any other model) has a
@@
         cutpoint-specific effects.
+    \end{itemize}
   \item Model-free methods don't define a common scale for comparing
@@
     measures can be \emph{nontransitive}, so it's not clear how to
-    use them in $\geq 2$-arm studies
-  \item For the same reason, model-based methods have additional
-    advantages, for example
-    the possibility of adjustment for covariates.
+    use them in $\geq 2$-arm studies.
+
+    That said, when comparing more than two distributions \emph{that
+    can be ordered} (i.e. when $x$ as well as $y$ is ordinal),
+    Somers' $D$ is a generalization of the win difference, Harrell's
+    $c$-index is a generalization of the win probability, and
+    \citet{newsonParametersNonparametricStatistics2002} discusses a
+    generalization of the win ratio.
+
+  \item Model-based methods allow one to adjust for covariates.
 
@@
     be fleshed out.)
+  \item This may unintentionally make interpreting estimated effect
+    sizes for the purposes of both study planning and interpreting results.
 \end{itemize}
@@
 Simulations show a typically very strong association between
-proportional odds ratios and $c$-index
+[double check] log common odds ratios and log win ratios (both of
+which generalize the odds ratio for a binary variable)
 \citep{harrellViolationProportionalOdds2020}.
 
-We should distinguish between effect measures used for \emph{analysis} and for
-\emph{reporting}. It may make sense to report the probabilistic index or a
+We should distinguish between effect measures used for \emph{design
+and decision-making},
+for \emph{analysis}, and for
+\emph{reporting}. For example it might make most sense to design a
+study and define decision rules using a common odds
+ratio/proportional odds model, but to
+additionally report results using the win probability..
+It may make
+sense to report the probabilistic index or a
 single-cutpoint binary outcome even if the analysis is undertaken using a
-proportional odds model.
+proportional odds model. Reporting win statistics based on model fits is
+suggested in \todo{one of}
+\cite{agrestiOrdinalProbabilityEffect2017,agrestiSimpleWaysInterpret2018}
+
+\section{Conclusion}
 

### 2025-06-06 constructing_ordinal_outcomes.tex
Bunch of work

@@
 
+\begin{abstract}
+  Ordinal outcomes or endpoints are an increasingly popular choice in
+  clinical trials.
+  They have been recently popular in the study of COVID-19 and in
+  adaptive platform
+  trials.
+  This paper is a guide to techniques for constructing ordinal
+  outcomes, aimed at clinicians and their statistical collaborators.
+  We categorize and illustrate ways combining existing binary,
+  ordinal, or continuous
+  outcomes through three broad mechanisms: sequential composition
+  (overriding), hierarchical composition (tie-breaking), and parallel
+  composition.
+  We discuss how this fine-graining impacts validity and efficiency, and also
+  suggest reasons why coarser outcomes might be preferred.
+  We also situate ``generalized pairwise comparisons'' and
+  ``prioritized outcomes'' approaches, like the win
+  ratio and desirability of outcome
+  ranking (DOOR) methodologies, in the broader context of ordinal
+  outcomes, suggesting that methods and approaches from each field
+  might be useful in the other.
+\end{abstract}
+
 \tableofcontents
@@
 
-Ordinal outcomes have become increasingly prominent in clinical
+Ordinal outcomes (or ``endpoints'') have become increasingly
+prominent in clinical
 resarch, particularly following the COVID-19 pandemic, when ordinal
@@
 
+Ordinal outcomes rank patient outcomes from better to worse, but do
+not compare \emph{differences} in patient outcomes. Thus binary
+outcomes are ordinal, and so are continuous outcomes (though
+``treating them ordinally'' means ignoring numeric values).
+
 Some advantages of ordinal outcomes are that they can
 \begin{enumerate}
-  \item Provide more information that binary outcomes, including
-    composite outcomes, making study designs more sensitive and precise, and
+  \item Provide more information than binary outcomes, making study
+    designs more sensitive and precise, and
     therefore more efficient and ethical
@@
 The goal of this manuscript is to serve as a guide for clinicians and trialists
-in developing ordinal outcomes for their studies. We consider both the outcome
-itself (the data recorded on each patient) and effect measures or ``endpoints''
-(the metric used to compare distributions of the outcome across patient
-groups).
+in developing ordinal outcomes for their studies.
 
-We want to distinguish \emph{choosing} or \emph{constructing} an outcome/effect
-measure from \emph{validating} the outcome/effect measure using external data.
-This paper is about the first task, choosing (or constructing) the outcome and
-effect measure. In practice choosing an outcome and effect measure and
-validating that outcome and effect measure in pre-trial data will be an
-iterative process. Our focus in this paper is on the ``positive'' side of the
-process--namely coming up with, and tweaking, proposed outcomes and effect
-measures--rather than the ``negative'' side, namely testing or checking
-proposals with data to assess whether they have desired properties.
+It is important to distinguish ``outcomes'' and ``effect measures'' -- both
+sometimes referred to under the banner of ``endpoints''. Here we
+focus on outcomes; we discuss endpoints in a separate paper \todo{cite}.
 
-A final goal of this paper is to situate the literature around ``generalized
+Another goal is to situate the literature around ``generalized
 pairwise comparisons of prioritized outcomes''
@@
 
-\subsection{Constructing versus validating}
+\subsection{The iterative process of constructing and validating}
+
+We want to distinguish \emph{choosing} or \emph{constructing} an outcome/effect
+measure from \emph{validating} the outcome/effect measure using external data.
+This paper is about the first task, choosing (or constructing) the outcome and
+effect measure. In practice choosing an outcome and effect measure and
+validating that outcome and effect measure in pre-trial data will be an
+iterative process. Our focus in this paper is on the ``positive'' side of the
+process--namely coming up with, and tweaking, proposed outcomes and effect
+measures--rather than the ``negative'' side, namely testing or checking
+proposals with data to assess whether they have desired properties.
+
+\section{Reasons to combine (or fine-grain) ordinal outcomes}
+
+\subsection{Capturing a broader spectrum of benefit and harm}
+
+Example: PASSPORT outcome (Figure \ref{fig:passport_outcome}),
+designed to capture benefit and harm across sepsis patients with a
+range of different disease severities.
+
+\subsection{Incorporating intercurrent events, especially death,
+for validity}
+
+The intercurrent event could be (1) incorporated as a level (2) used
+for tie-breaking or overriding.
+
+This is a version of the ``composite strategy'' as described in e.g.
+\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
+more efficiency than creating a \emph{binary} composite outcome.
+
+For example overriding a numeric outcome by death.
+Example: REMAP-CAP: organ-free survival, overridden by death
+(Figure \ref{fig:remap_cap_override}).
+
+\begin{figure}
+  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
+  \caption{\textbf{Overriding a continuous outcome, illustrated by
+    REMAP-CAP}. Outcome from the REMAP-CAP adaptive
+    platform trial
+    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
+    Illustrates overriding a numeric ordinal outcome (organ
+    support-free days) by a binary ordinal outcome (death; this is
+    ``sequential composition''), which
+    would also be an intercurrent event. This is another example of
+  sequential composition.}
+  \label{fig:remap_cap_override}
+\end{figure}
+
+\subsection{Increasing efficiency by shrinking the largest category}
+
+Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
+efficiency (variance) of an unadjusted treatment comparison using an ordinal
+versus continuous outcome in the proportional odds model is given by $1 - \sum
+\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
+over control and experimental groups) of outcome category $c$.
+Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
+categories when the probabilities are all equal: $\overline p_c = 1/C$.
+
+The results are based on estimating a common odds ratio, but in large
+enough samples are equivalent to the Mann-Whitney-Wilcoxon test, and
+therefore to testing the various ``win nulls'' (win probability is
+  0.5, win difference
+is 0, win ratio is 1).
+
+We have simulation results and some maths to show that the value of $\sum_c
+\overline p_c^3$ -- and therefore the (relative) efficiency of the
+ordinal outcome -- is essentially a function of the largest category
+probability $\overline{p}_{\max} = \max \overline{p}_c$. See Figure
+\ref{fig:p_max} for an illustration of the maths.
+
+\begin{figure}
+  \includegraphics[width=6in]{p_max_controls_efficiency.png}
+  \caption{\textbf{Efficiency gains are controlled by the probability
+    of the largest category.} Bounds on the relative efficiency of an
+    ordinal outcome
+    compared to its continuous counterpart as a function of the maximum
+    category probability $p_\text{max}$. Results are based on
+    estimating a common odds ratio or testing stochastic ordering
+    using the Mann-Whitney-Wilcoxon test. The upper bound (best case
+    relative efficiency for a given $p_\text{max}$) is always achieved in the
+    limit of infinitely many categories ($C = \infty$). The lower bound
+    is always achieved when the number of categories is as small as
+    possible for a given $p_\text{max}$; in particular when
+  $p_\text{max} \geq 0.5$ the worst case occurs with 2 categories.}
+  \label{fig:p_max}
+\end{figure}
+
+Thus simple advice is: for statistical efficiency, try to make the proportion
+of patients in the largest category small. However a converse is: if the
+proportion of patients in the largest category is small, not much is gained by
+fine-graining; a coarser outcome might be easier to communicate.
+
+\subsection{Augmenting a state or time-to-event outcome with
+longitudinal information}
+
+Distinguish between ``state'' and ``trajectory'' outcomes. State
+variables summarize patient outcomes at a particular moment in time.
+Trajectory variables, including times-to-event, summarize patient
+state over a timecourse. (In practice ``state'' variables are
+actually summaries over an interval like ``day 28 post-randomization''.)
+
+Trajectory and state variables can be incorporated into an ordinal outcome. For
+example (from PASSPORT; Figure \ref{fig:passport_outcome}) ``worst pSOFA while
+hospitalized until day 28'' and ``length of hospitalization'', which are
+longitudinal variables (summary of a trajectory), are combined with the day 28
+state variable ``death''. The REMAP-CAP outcome is similar (Figure
+\ref{fig:remap_cap_override}).
+
+Most time-to-event analyses are in fact analyses of an ordinal outcome. The Cox
+model for time-to-event data treats time as ordinal: because the baseline
+hazard is not modelled, the model itself is invariant to monotone rescalings of
+the time axis, and the (partial) likelihood (ignoring ties) depends only on the
+ranks of the data (i.e., is a marginal likelihood). The Cox model is
+essentially an ordinal stopping ratio model with complementary log-log link.
+(Plus handling right-censoring.)
+
+An important concern when incorporating longitudinal information is
+to avoid including nonclinical effects of treatment/exposure in the
+outcome; \citet{ongUnlockingDOORHow2023} discuss/criticize RADAR
+methods from this perspective.
+
+Win methods were inaugurated in
+\citet{finkelsteinCombiningMortalityLongitudinal1999} with the
+express motivation of combining longitudinal and time-to-event information.
 
@@
     \emph{overriding events} when $A$ is binary.
+
   \item \textbf{Hierarchical composition} Complete tie-breaking of $A$ by
-    $B$ (``prioritized outcomes''; lexicographic order) within
+    $B$ (``prioritized outcomes''; lexicographic ordering) within
     \emph{every} level of $A$.
@@
     This is what win methods do.
+
+    Sequential composition is complete-tie breaking followed by coarsening.
+
   \item \textbf{Parallel composition} This is where levels of $A$ are
@@
     This is not so precisely determined as the other two. We need to
-    decided \emph{how} to aggregate.
+    decided \emph{how} to aggregate. Some ways are
+    \begin{itemize}
+      \item summing up numeric
+        scores
+      \item counting different kinds of events
+      \item taking the worst level across ordinal outcomes considered
+        in parallel (as in BANDICOOT; Figure \ref{fig:parallel_bandicoot}).
+    \end{itemize}
 \end{enumerate}
@@
     composition/overriding,
-  \item "$/$" to denote hierarchical composition/complete tiebreaking
+  \item "$/$" to denote hierarchical composition/tiebreaking
     (note that this is an asymmetric operation), and
@@
 Aside: \citet{wittkowskiCombiningSeveralOrdinal2004} consider combining
-ordinal variables into partial orders. This makes more sense in the
-generalized pairwise comparisons framework.
+ordinal variables into \emph{partial} orders. Beyond the scope of this paper.
 
@@
   % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
-  \caption{Ordinal outcome used in the planned PASSPORT adaptive
+  \caption{\textbf{Sequential composition (overriding) illustrated
+    with the PASSPORT outcome} Ordinal outcome used in the planned
+    PASSPORT adaptive
     platform trial. This illustrates sequential composition/overriding:
@@
   \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
-  \caption{``Angina Symptom Score'' ordinal outocome develped in
+  \caption{\textbf{Hierarchical composition (complete tie-breaking)
+      illustrated with the
+    the ORBITA-2 outcome.} ``Angina
+    Symptom Score'' ordinal outocome develped in
     \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
-    (1) complete tie-breaking (of units of antianginal medication by
-    episodes of angina) and (2) overriding (of angina episodes and
+    (1) hierarchical composition or complete tie-breaking (of units
+      of antianginal medication by
+    episodes of angina) and also (2) overriding (of angina episodes and
   angina medication by adverse clinical events)}
@@
 
-\begin{figure}
-  \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
+\begin{figure} \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
   % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
-  \caption{Ordinal outcome used in the planned BANDICOOT adaptive
-    platform trial \citep{walkerCodesigningNovelOrdinal2025}.
-  Illustrates \textbf{parallel composition} of component outcomes.}
-  \label{fig:parallel_bandicoot}
+  \caption{\textbf{Parallel composition illustrated with the BANDICOOT outcome}.
+    Ordinal outcome used in the planned BANDICOOT adaptive platform trial
+    \citep{walkerCodesigningNovelOrdinal2025}. Illustrates \textbf{parallel
+  composition} of component outcomes.} \label{fig:parallel_bandicoot}
 \end{figure}
 
-\subsection{Converting back to numeric scores}
+\subsubsection{Parallel composition by combining numeric scores}
 
-Treat each input factor as
-numeric then sum them up. I believe \citet{ongUnlockingDOORHow2023}
-has examples?
+One option: convert each input factor to a numeric value then sum them up.
+
+A specific example of this, described e.g. by \citet{ongUnlockingDOORHow2023} in
+the infectious disease context, is counting events of a particular
+type (corresponds to converting the presence/absence of the event to
+a 1/0 binary variable).
 
@@
 
-\subsection{Pairwise comparison methods (win, DOOR) implicitly construct ordinal
-outcomes}
+\subsection{Pairwise comparison methods (win and DOOR) implicitly construct
+ordinal outcomes}
 
@@
     hierarchical composition (complete tie-breaking/prioritized outcomes)
-  \item Targeting a nonparametric effect measure defined via the
-    ``pairs'' device (win probability/win ratio/win odds/etc.)
-  \item Estimating that effect measure in a nonparametric/flexible
-    manner using pairs, often also with devices for accounting for censoring
+  \item
+    Targeting a nonparametric effect measure defined via the
+    ``pairs'' device (win
+    probability/win ratio/win odds/etc.)
+  \item Estimating that effect measure in a
+    nonparametric/flexible manner using pairs, often also with devices for
+    accounting for censoring
 \end{enumerate}
@@
   \item Win methods can make use of
-    plots and diagnostics for general ordinal outcomes, for the
-    \emph{implicit} ordinal variable
+    plots and diagnostics for general ordinal outcomes, for the \emph{implicit}
+    ordinal variable
     \begin{itemize}
-      \item
-        comparative CDF plots for the
+      \item comparative CDF or ``survival''
+        (complementary CDF) plots for assessing stochastic
+        monotonicity \todo{this
+          seems not done routinely? perhaps because of the difficulty
+        of incorporating censoring?}
       \item ridit plots
-        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984}
-        relative to a reference group, which should allow one to
-        display both full eCDF information and also win statistics
-        (based on the interpretation of the mean ridit as a win probability)
+        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984} relative to a
+        reference group, which should allow one to display both full
+        eCDF information
+        and also win statistics (based on the interpretation of the
+          mean ridit as a win
+        probability)
     \end{itemize}e.g.
-  \item Ordinal regression models can use win methods for model
-    checking; it is
-    also possible to present a model-based win effect measure
-    alongside a common
-    odds ratio for example. See
-    \citep{agrestiOrdinalProbabilityEffect2017} for example, which
-    covers both latent variable and predicted probability versions.
-  \item The outcomes \emph{implicitly} constructed using
-    DOOR and win methods may contain many levels, and may contain
-    levels where an
-    effect is not so meaningful (e.g. time-to-death when the followup time is
-    short). By \emph{explicitly} constructing the associated
-    ordinal outcome one
-    might be forced to be more aware of this
-  \item What is a reasonable win
-    analysis of a specific component score? From the traditional
-    ordinal endpoint
-    school of thought what would make most sense is to \emph{only
-      tie-break for the
-    lowest and highest levels of that score}, and e.g. plot eCDFs of this new
-    variable. Is this what the plots do that break down wins and losses by
-    component score?
+  \item Ordinal regression models can use win
+    methods for model checking; it is also possible to present a model-based win
+    effect measure alongside a common odds ratio for example. See
+    \citep{agrestiOrdinalProbabilityEffect2017} for example, which covers both
+    latent variable and predicted probability versions.
+  \item The outcomes
+    \emph{implicitly} constructed using DOOR and win methods may contain many
+    levels, and may contain levels where an effect is not so meaningful (e.g.
+    time-to-death when the followup time is short). By \emph{explicitly}
+    constructing the associated ordinal outcome one might be forced to be more
+    aware of this
+  \item What is a reasonable win analysis of a specific component
+    score? From the traditional ordinal endpoint school of thought
+    what would make
+    most sense is to \emph{only tie-break for the lowest and highest
+      levels of that
+    score}, and e.g. plot eCDFs of this new variable. Is this what the plots do
+    that break down wins and losses by component score?
 \end{itemize}
@@
 
-\section{Reasons to combine ordinal outcomes}
-
-\subsection{Providing a more holistic measure of benefit and harm}
-
-\subsection{Incorporating intercurrent events, especially death,
-for validity}
-
-The intercurrent event could be (1) incorporated as a level (2) used
-for tie-breaking or overriding.
-
-This is a version of the ``composite strategy'' as described in e.g.
-\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
-more efficiency than creating a \emph{binary} composite outcome.
-
-For example overriding a numeric outcome by death.
-Example: REMAP-CAP: organ-free survival, overridden by death
-(Figure \ref{fig:remap_cap_override}).
-
-\begin{figure}
-  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
-  \caption{Ordinal outcome used in the REMAP-CAP adaptive
-    platform trial
-    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
-    Illustrates overriding a numeric ordinal outcome (organ
-    support-free days) by a binary ordinal outcome (death; this is
-    ``sequential composition''), which
-    would also be an intercurrent event. This is another example of
-  sequential composition.}
-  \label{fig:remap_cap_override}
-\end{figure}
-
-\subsection{Increasing precision by reducing the probability of the
-largest category}
-
-Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
-efficiency (variance) of an unadjusted treatment comparison using an ordinal
-versus continuous outcome in the proportional odds model is given by $1 - \sum
-\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
-over control and experimental groups) of outcome category $c$.
-
-Same should apply to the Mann-Whitney-Wilcoxon test, and therefore to
-the win probability, and therefore at least approximately to win
-difference/win ratio/win odds.
-
-Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
-categories when the probabilities are all equal: $\overline p_c = 1/C$.
-
-We have simulation results and some maths to show that the value of $\sum_c
-\overline p_c^3$ is \emph{essentially a function of the largest category
-probability} $\overline{p_{\max}} = \max \overline{p_c}$. See Figure
-\ref{fig:p_max} for an illustration of the maths.
-
-\begin{figure}
-  \includegraphics[width=6in]{p_max_controls_efficiency.png}
-  \caption{Bounds on the relative efficiency of an ordinal outcome
-    compared to its continuous counterpart as a function of the maximum
-    category probability $p_\text{max}$. The upper bound (best case
-    relative efficiency for a given $p_\text{max}$) is always achieved in the
-    limit of infinitely many categories ($C = \infty$). The lower bound
-    is always achieved when the number of categories is as small as
-  possible for a given $p_\text{max}$.}
-  \label{fig:p_max}
-\end{figure}
-
-Thus simple advice is: for statistical efficiency, try to make the proportion
-of patients in the largest category small.
-
-However a converse is: if the proportion of patients in the largest
-category is
-small, not much is gained by tie-breaking.
-
-\subsection{Augmenting a state outcome with longitudinal
-information}
-
-Distinguish between ``state'' and ``trajectory'' variables. State
-variables summarize patient outcomes at a particular moment in time.
-Trajectory variables, including times-to-event, summarize patient
-state over a timecourse.
-
-Trajectory and state variables can be incorporated into an ordinal outcome. For
-example (from PASSPORT) ``worst pSOFA while hospitalized until day 28'' and
-``length of hospitalization'', which are longitudinal variables (summary of a
-trajectory), are combined with the day 28 state variable ``death''.
-
-Incorporating trajectory variables into an ordinal outcome may seem unnatural.
-But arguably most time-to-event analyses are in fact analyses of an
-ordinal outcome. The Cox model for time-to-event data treats time as ordinal:
-because (1) the baseline hazard is not modelled, (2) the model itself is
-invariant to monotone rescalings of the time axis, and (3) the (partial)
-likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
-marginal likelihood). The Cox model is essentially an ordinal
-stopping ratio model with complementary log-log link. (Perhaps a
-difference is censoring.)
-
 \section{Reasons to coarsen ordinal outcomes}
 
-\subsection{Coarsening to avoid directly comparing levels}
+\subsection{Coarsening to avoid comparing ambiguous levels (validity)}
 
@@
 
-(It isn't really a problem for model fitting with modern software.)
+(Outocmes with very many levels -- even up to of the order of 100 --
+aren't really a problem for model fitting with modern software.)
 

### 2025-06-06 constructing_ordinal_outcomes.tex
Follman et al 2020 new outcome approach

@@
 
+    This is what win methods do \todo{I \emph{THINK}. Need to go
+      through an example very carefully to check it's not in fact
+      sequential composition. Check
+      \citep{follmannAnalysisOrderedComposite2020} and the $x$ axis of
+    their survival plots}
   \item \textbf{Hierarchical composition} Complete tie-breaking of $A$ by
@@
     \emph{every} level of $A$.
-
-    This is what win methods do.
 
@@
   \item Estimating that effect measure in a
-    nonparametric/flexible manner using pairs, often also with devices for
-    accounting for censoring
+    nonparametric/flexible manner using pairs, \textbf{with devices for
+    accounting for censoring}
 \end{enumerate}
@@
 This paper is focused on the first of these steps only.
+
+\citet{follmannAnalysisOrderedComposite2020} is an example of the
+approach of explicitly constructing a new ordinal outcome based on a
+hierarchy (in this case a hierarchy of 3 times-to-event, with event
+times for worse events overriding event times for better ones.).
 
@@
         (complementary CDF) plots for assessing stochastic
-        monotonicity \todo{this
-          seems not done routinely? perhaps because of the difficulty
-        of incorporating censoring?}
+        monotonicity. \citet{follmannAnalysisOrderedComposite2020}
+        also does this.
       \item ridit plots

### 2025-06-19 constructing_ordinal_outcomes.tex
Many updates; start to restructure initial sections

@@
 \begin{abstract}
-  Ordinal outcomes or endpoints are an increasingly popular choice in
+  Ordinal outcomes (or ``endpoints'') are an increasingly popular choice in
   clinical trials.
-  They have been recently popular in the study of COVID-19 and in
-  adaptive platform
-  trials.
   This paper is a guide to techniques for constructing ordinal
   outcomes, aimed at clinicians and their statistical collaborators.
-  We categorize and illustrate ways combining existing binary,
+  We categorize and illustrate ways of combining existing binary,
   ordinal, or continuous
@@
 
+\subsection{Ordinal outcomes and their advantages}
+
 Ordinal outcomes (or ``endpoints'') have become increasingly
@@
 scales developed by the WHO and NIAID were used in prominent trials
-of drugs repurposed to treat the disease \todo{cite}. They have also
-been a popular choice in adaptive platform trials \todo{cite}.
+\citep{beigelRemdesivirTreatmentCovid192020,
+marshallMinimalCommonOutcome2020}. They have also
+been used in a number of adaptive platform trials
+\citep{angusREMAPCAPRandomizedEmbedded2020, walkerCodesigningNovelOrdinal2025}.
 
-Ordinal outcomes rank patient outcomes from better to worse, but do
-not compare \emph{differences} in patient outcomes. Thus binary
-outcomes are ordinal, and so are continuous outcomes (though
-``treating them ordinally'' means ignoring numeric values).
+Ordinal outcomes rank patient outcomes from better to worse, but
+\emph{differences} in patient outcomes are
+not treated as meaningful. Binary
+outcomes are ordinal -- they are the simplest kind of
+ordinal outcome -- and so too are continuous outcomes, though
+``treating them ordinally'' involves ignoring their numeric values.
 
-Some advantages of ordinal outcomes are that they can
-\begin{enumerate}
-  \item Provide more information than binary outcomes, making study
-    designs more sensitive and precise, and
-    therefore more efficient and ethical
+Two desirable features of outcomes in general are \emph{validity} and
+\emph{efficiency}.
 
+\textbf{Valid} outcomes reflect
+patient benefit and harm: better/worse outcomes are judged respectively
+better/worse by patients, are objectively respectively better/worse for
+patients, or are reliable surrogates/predictors of outcomes that are
+respectively better/worse for patients.
+
+\begin{itemize}
   \item Allow for the incorporation of intercurrent events into one's primary
-    outcome, avoiding complexities of alternative estimand strategies like
-    principal stratum estimands (particular for truncating events
-    like death) while
-    yielding more statistical efficiency than traditional binary
-    composite outcomes
+    outcome, avoiding complexities of alternative strategies like
+    principal stratum
+    estimands (particular for truncating events like death) while yielding more
+    statistical efficiency than traditional binary composite outcomes
 
@@
     ``state'' outcomes in a single measure.
+\end{itemize}
 
-\end{enumerate}
+\textbf{Efficient} outcomes
+capture sufficient information to discriminate among better- and worse-off
+patients without large sample sizes.
 
-The goal of this manuscript is to serve as a guide for clinicians and trialists
+\begin{itemize}
+  \item Provide more information than binary outcomes, making study
+    designs more sensitive and precise, and
+    therefore more efficient and ethical
+\end{itemize}
+
+For the purposes of this
+article, set aside ease of collection, low missingess, objective
+ascertainability, etc.
+
+A key feature of ordinal outcomes is the ease with which they can be
+\emph{combined}.
+
+\subsection{Desiderata for (ordinal) outcomes: validity and efficiency}
+
+We begin with some general desiderata for outcomes and effect measures in
+clinical studies.
+
+An example of an outcome which is valid but inefficient might be
+death as a binary outcome in a context where deaths
+are rare.
+
+An example of an outcome which is efficient but which may be invalid
+might be continuously-measured concentrations of a biomarker which is a poor
+surrogate for disease progression.
+
+\subsection{The iterative process of constructing and validating}
+
+We want to distinguish \emph{choosing} or \emph{constructing} an
+outcome/effect
+measure from \emph{validating} the outcome/effect measure using
+external data.
+This paper is about the first task, choosing (or constructing)
+the outcome and
+effect measure. In practice choosing an outcome and effect measure and
+validating that outcome and effect measure in pre-trial data will be an
+iterative process. Our focus in this paper is on the ``positive''
+side of the
+process--namely coming up with, and tweaking, proposed outcomes and effect
+measures--rather than the ``negative'' side, namely testing or checking
+proposals with data to assess whether they have desired properties.
+
+\subsection{Objectives/contributions}
+
+The goal of this manuscript is to serve as a guide for clinicians
+and trialists
 in developing ordinal outcomes for their studies.
-
-It is important to distinguish ``outcomes'' and ``effect measures'' -- both
-sometimes referred to under the banner of ``endpoints''. Here we
-focus on outcomes; we discuss endpoints in a separate paper \todo{cite}.
 
@@
 
-\subsection{Desiderata for (ordinal) outcomes: validity and efficiency}
-
-We begin with some general desiderata for outcomes and effect measures in
-clinical studies.
-
-We want outcomes that
-\begin{enumerate}
-  \item are \textbf{valid} in that they reflect patient benefit and harm:
-    better/worse outcomes are judged respectively better/worse by patients, are
-    objectively respectively better/worse for patients, or are reliable
-    surrogates/predictors of outcomes that are respectively better/worse for
-    patients
-  \item are \textbf{efficient} in the sense of capturing
-    sufficient information to
-    discriminate among better- and worse-off patients without large
-    sample sizes.
-\end{enumerate}
-(For the purposes of this article, set aside ease of collection, low
-missingess, objective ascertainability, etc.)
-
-An example of an outcome which is valid but inefficient might be
-death as a binary outcome in a context where deaths
-are rare.
-
-An example of an outcome which is efficient but which may be invalid
-might be continuously-measured concentrations of a biomarker which is a poor
-surrogate for disease progression.
-
-One point of view is that one would ``ideally'' analyze ordinal outcomes by
-reporting the associated expected utilities \todo{cite Berry talk; example of
-utility-weighted Rankin scale} and that since existing ordinal methods supply
-such utilities implicitly why not just supply them explicitly. A counterpoint
-is that the methods do not, in fact, supply utilities implicitly.
-Another is that ordinal outcomes let one ``hedge'' about precise
-utillities and this is OK.
-
-\subsection{The iterative process of constructing and validating}
-
-We want to distinguish \emph{choosing} or \emph{constructing} an outcome/effect
-measure from \emph{validating} the outcome/effect measure using external data.
-This paper is about the first task, choosing (or constructing) the outcome and
-effect measure. In practice choosing an outcome and effect measure and
-validating that outcome and effect measure in pre-trial data will be an
-iterative process. Our focus in this paper is on the ``positive'' side of the
-process--namely coming up with, and tweaking, proposed outcomes and effect
-measures--rather than the ``negative'' side, namely testing or checking
-proposals with data to assess whether they have desired properties.
+It is important to distinguish ``outcomes'' and ``effect measures'' -- both
+sometimes referred to under the banner of ``endpoints''. Here we
+focus on outcomes; we discuss endpoints in a separate paper \todo{cite}.
 
@@
 efficiency (variance) of an unadjusted treatment comparison using an ordinal
-versus continuous outcome in the proportional odds model is given by $1 - \sum
+versus continuous outcome in the proportional odds model is given
+by $1 - \sum
 \overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
@@
     using the Mann-Whitney-Wilcoxon test. The upper bound (best case
-    relative efficiency for a given $p_\text{max}$) is always achieved in the
+    relative efficiency for a given $p_\text{max}$) is always
+    achieved in the
     limit of infinitely many categories ($C = \infty$). The lower bound
@@
 
-Thus simple advice is: for statistical efficiency, try to make the proportion
+Thus simple advice is: for statistical efficiency, try to make
+the proportion
 of patients in the largest category small. However a converse is: if the
-proportion of patients in the largest category is small, not much is gained by
+proportion of patients in the largest category is small, not much
+is gained by
 fine-graining; a coarser outcome might be easier to communicate.
@@
 
-Trajectory and state variables can be incorporated into an ordinal outcome. For
-example (from PASSPORT; Figure \ref{fig:passport_outcome}) ``worst pSOFA while
+Trajectory and state variables can be incorporated into an
+ordinal outcome. For
+example (from PASSPORT; Figure \ref{fig:passport_outcome})
+``worst pSOFA while
 hospitalized until day 28'' and ``length of hospitalization'', which are
-longitudinal variables (summary of a trajectory), are combined with the day 28
+longitudinal variables (summary of a trajectory), are combined
+with the day 28
 state variable ``death''. The REMAP-CAP outcome is similar (Figure
@@
 
-Most time-to-event analyses are in fact analyses of an ordinal outcome. The Cox
+Most time-to-event analyses are in fact analyses of an ordinal
+outcome. The Cox
 model for time-to-event data treats time as ordinal: because the baseline
-hazard is not modelled, the model itself is invariant to monotone rescalings of
-the time axis, and the (partial) likelihood (ignoring ties) depends only on the
+hazard is not modelled, the model itself is invariant to monotone
+rescalings of
+the time axis, and the (partial) likelihood (ignoring ties)
+depends only on the
 ranks of the data (i.e., is a marginal likelihood). The Cox model is
@@
     \ref{fig:passport_outcome}, is given by
-    \[ \text{day 28 status} \leftarrow \text{max pSOFA} \leftarrow \text{days
+    \[ \text{day 28 status} \leftarrow \text{max pSOFA}
+      \leftarrow \text{days
       hospitalized}
@@
     \textbf{Hierarchical composition} ``Angina Symptom Score'' outcome from
-    \citet{rajkumarPlaceboControlledTrialPercutaneous2023}, depicted in Figure
+    \citet{rajkumarPlaceboControlledTrialPercutaneous2023},
+    depicted in Figure
     \ref{fig:rajkumar_et_al_outcome}, is given by \[
@@
     \[
-      (\text{death, graft loss, or relapse}) \leftarrow (\text{viraemia} \;|\;
+      (\text{death, graft loss, or relapse}) \leftarrow
+      (\text{viraemia} \;|\;
       \text{CD4+ count} \;|\; \text{organ support})
@@
 Aside: \citet{wittkowskiCombiningSeveralOrdinal2004} consider combining
-ordinal variables into \emph{partial} orders. Beyond the scope of this paper.
+ordinal variables into \emph{partial} orders. Beyond the scope of
+this paper.
 
@@
     the ORBITA-2 outcome.} ``Angina
-    Symptom Score'' ordinal outocome develped in
+    Symptom Score'' ordinal outcome develped in
     \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
@@
 
-\begin{figure} \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
+\begin{figure}
+  \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
   % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
-  \caption{\textbf{Parallel composition illustrated with the BANDICOOT outcome}.
+  \caption{\textbf{Parallel composition illustrated with the
+    BANDICOOT outcome}.
     Ordinal outcome used in the planned BANDICOOT adaptive platform trial
@@
 
-A specific example of this, described e.g. by \citet{ongUnlockingDOORHow2023} in
+A specific example of this, described e.g. by
+\citet{ongUnlockingDOORHow2023} in
 the infectious disease context, is counting events of a particular
@@
   \item Win methods can make use of
-    plots and diagnostics for general ordinal outcomes, for the \emph{implicit}
+    plots and diagnostics for general ordinal outcomes, for the
+    \emph{implicit}
     ordinal variable
@@
       \item ridit plots
-        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984} relative to a
+        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984}
+        relative to a
         reference group, which should allow one to display both full
@@
   \item Ordinal regression models can use win
-    methods for model checking; it is also possible to present a model-based win
+    methods for model checking; it is also possible to present a
+    model-based win
     effect measure alongside a common odds ratio for example. See
-    \citep{agrestiOrdinalProbabilityEffect2017} for example, which covers both
+    \citep{agrestiOrdinalProbabilityEffect2017} for example,
+    which covers both
     latent variable and predicted probability versions.
   \item The outcomes
-    \emph{implicitly} constructed using DOOR and win methods may contain many
-    levels, and may contain levels where an effect is not so meaningful (e.g.
+    \emph{implicitly} constructed using DOOR and win methods may
+    contain many
+    levels, and may contain levels where an effect is not so
+    meaningful (e.g.
     time-to-death when the followup time is short). By \emph{explicitly}
-    constructing the associated ordinal outcome one might be forced to be more
+    constructing the associated ordinal outcome one might be
+    forced to be more
     aware of this
@@
       levels of that
-    score}, and e.g. plot eCDFs of this new variable. Is this what the plots do
+    score}, and e.g. plot eCDFs of this new variable. Is this
+    what the plots do
     that break down wins and losses by component score?
@@
 
-\section{Conclusions}
+\section{Discussion and Conclusions}
+
+One point of view is that one would ``ideally'' analyze ordinal outcomes by
+reporting the associated expected utilities \todo{cite Berry
+  talk; example of
+utility-weighted Rankin scale} and that since existing ordinal
+methods supply
+such utilities implicitly why not just supply them explicitly. A
+counterpoint
+is that the methods do not, in fact, supply utilities implicitly.
+Another is that ordinal outcomes let one ``hedge'' about precise
+utillities and this is OK.
 

### 2025-06-19 constructing_ordinal_outcomes.tex
Attempt to tighten up intro w Kate+Chris feedback

@@
 \begin{abstract}
-  Ordinal outcomes (or ``endpoints'') are an increasingly popular choice in
+  Ordinal outcomes are an increasingly popular choice for
   clinical trials.
   This paper is a guide to techniques for constructing ordinal
-  outcomes, aimed at clinicians and their statistical collaborators.
-  We categorize and illustrate ways of combining existing binary,
+  outcomes out of pre-existing components, aimed at clinicians and their
+  statistical collaborators.
+  We discuss three broad mechanisms for combining existing binary,
   ordinal, or continuous
-  outcomes through three broad mechanisms: sequential composition
-  (overriding), hierarchical composition (tie-breaking), and parallel
-  composition.
-  We discuss how this fine-graining impacts validity and efficiency, and also
-  suggest reasons why coarser outcomes might be preferred.
-  We also situate ``generalized pairwise comparisons'' and
-  ``prioritized outcomes'' approaches, like the win
-  ratio and desirability of outcome
-  ranking (DOOR) methodologies, in the broader context of ordinal
-  outcomes, suggesting that methods and approaches from each field
-  might be useful in the other.
+  outcomes into an ordinal composite---overriding (or
+  ``sequential composition''), tie-breaking (or ``hierarchical
+  composition''), and what we term ``parallel
+  composition''---and illustrate each mechanism with a real-world example.
+  We discuss how these mechanisms can improve the validity and
+  efficiency of ordinal outcomes, but also why coarsening outcomes by
+  combining levels might be preferred in practice.
+  Our approach is closely related to the literature on prioritized
+  outcomes (like win methods and
+    desirability of outcome
+  ranking methodologies), but we focus on the task of
+  constructing ordinal
+  outcomes and distinguish this from the related tasks of choosing
+  effect measures and estimators and
+  handling censoring.
 \end{abstract}
-
+\newpage
 \tableofcontents
@@
 
-\subsection{Ordinal outcomes and their advantages}
+\subsection{Ordinal outcomes, their recent popularity and advantages}
 
@@
 \citep{beigelRemdesivirTreatmentCovid192020,
-marshallMinimalCommonOutcome2020}. They have also
+marshallMinimalCommonOutcome2020}. Ordinal outcomes have also
 been used in a number of adaptive platform trials
-\citep{angusREMAPCAPRandomizedEmbedded2020, walkerCodesigningNovelOrdinal2025}.
+\citep{angusREMAPCAPRandomizedEmbedded2020,
+walkerCodesigningNovelOrdinal2025}.
 
@@
 \emph{differences} in patient outcomes are
-not treated as meaningful. Binary
-outcomes are ordinal -- they are the simplest kind of
-ordinal outcome -- and so too are continuous outcomes, though
-``treating them ordinally'' involves ignoring their numeric values.
+not treated as meaningful.
 
@@
 \emph{efficiency}.
+\emph{Valid} outcomes reflect
+patient benefit and harm; \emph{efficient} outcomes
+are able to discriminate among better- and worse-off
+patients without large sample sizes.
 
-\textbf{Valid} outcomes reflect
-patient benefit and harm: better/worse outcomes are judged respectively
-better/worse by patients, are objectively respectively better/worse for
-patients, or are reliable surrogates/predictors of outcomes that are
-respectively better/worse for patients.
+A great strength of ordinal outcomes is the ease with which they can
+be combined.
+
+This can help with validity by
 
 \begin{itemize}
-  \item Allow for the incorporation of intercurrent events into one's primary
+  \item \emph{Handle intercurrent events} by incorporating them into a primary
     outcome, avoiding complexities of alternative strategies like
     principal stratum
-    estimands (particular for truncating events like death) while yielding more
-    statistical efficiency than traditional binary composite outcomes
+    estimands (particular for truncating events like death) while
+    yielding more
+    statistical efficiency than traditional binary composite outcomes.
 
-  \item Allow for the easy incorporation of safety events into a
-    primary outcome, accounting for many different aspects
-    of benefit/harm; at the same time, ordinal outcomes can be
-    analyzed without a
-    fully-specified utility function, which may require more
-    extensive development and validation or may vary widely across patients.
+  \item \emph{Jointly summarizing efficacy and safety} by incorporating
+    safety events into a
+    primary outcome.
 
-  \item Allow for the combination of discrete and continous outcome
+  \item \emph{Combining outcome measures of different types} Allow
+    for the combination of discrete and continous outcome
     measures (for
@@
 
-\textbf{Efficient} outcomes
-capture sufficient information to discriminate among better- and worse-off
-patients without large sample sizes.
+Moreover it can help with efficiency by
 
@@
 
-A key feature of ordinal outcomes is the ease with which they can be
-\emph{combined}.
-
-\subsection{Desiderata for (ordinal) outcomes: validity and efficiency}
-
-We begin with some general desiderata for outcomes and effect measures in
-clinical studies.
-
 An example of an outcome which is valid but inefficient might be
@@
 might be continuously-measured concentrations of a biomarker which is a poor
-surrogate for disease progression.
-
-\subsection{The iterative process of constructing and validating}
-
-We want to distinguish \emph{choosing} or \emph{constructing} an
-outcome/effect
-measure from \emph{validating} the outcome/effect measure using
-external data.
-This paper is about the first task, choosing (or constructing)
-the outcome and
-effect measure. In practice choosing an outcome and effect measure and
-validating that outcome and effect measure in pre-trial data will be an
-iterative process. Our focus in this paper is on the ``positive''
-side of the
-process--namely coming up with, and tweaking, proposed outcomes and effect
-measures--rather than the ``negative'' side, namely testing or checking
-proposals with data to assess whether they have desired properties.
+surrogate for disease progression, or a continuous outcome that
+excludes mortality.
 
@@
 
-It is important to distinguish ``outcomes'' and ``effect measures'' -- both
+\subsubsection{Non-objectives}
+
+\paragraph{Outcomes versus effect measures} It is important to
+distinguish ``outcomes'' and ``effect measures'' -- both
 sometimes referred to under the banner of ``endpoints''. Here we
 focus on outcomes; we discuss endpoints in a separate paper \todo{cite}.
+
+\paragraph{Constructing versus validating} We want to distinguish
+\emph{choosing} or \emph{constructing} an
+outcome/effect
+measure from \emph{validating} the outcome/effect measure using
+external data.
+This paper is about the first task, choosing (or constructing)
+the outcome and
+effect measure. In practice choosing an outcome and effect measure and
+validating that outcome and effect measure in pre-trial data will be an
+iterative process. Our focus in this paper is on the ``positive''
+side of the
+process--namely coming up with, and tweaking, proposed outcomes and effect
+measures--rather than the ``negative'' side, namely testing or checking
+proposals with data to assess whether they have desired properties.
 
@@
 longitudinal information}
+
+The process of combining ordinal outcomes is more flexible than it
+might seem, because many different kinds of outcomes can be treated as ordinal
+or summarized in ordinal form. Binary
+outcomes are ordinal -- they are the simplest kind of
+ordinal outcome -- and so too are continuous outcomes, though
+``treating them ordinally'' involves ignoring their numeric values.
+Longitudinal trajectory-outcomes can be summarized ordinally---for
+example ``worst measurement in the first 28 days of
+hospitalization''---and combined with ``state'' outcomes.
 
@@
   % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
-  \caption{\textbf{Sequential composition (overriding) illustrated
+  \caption{\textbf{Overriding (sequential composition) illustrated
     with the PASSPORT outcome} Ordinal outcome used in the planned
@@
   \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
-  \caption{\textbf{Hierarchical composition (complete tie-breaking)
+  \caption{\textbf{Tie-breaking (hierarchical composition)
       illustrated with the
@@
 Also, we "chunk" longer hospital stays more aggressively to avoid
-small categories, for interpretability's sake.
+small categories, for the sake of interpretability.
 
@@
 
+Our approach is closely related to that of
+\citet{follmannAnalysisOrderedComposite2020},
+who explicitly consider the ordinal outcome implicitly constructed by
+win methods from a time-to-event analysis of ordered events, and show
+how right-censoring of event times induces interval censoring of the
+cosntructed outcome.
+
 One point of view is that one would ``ideally'' analyze ordinal outcomes by
@@
 Another is that ordinal outcomes let one ``hedge'' about precise
-utillities and this is OK.
+utillities and this is OK:
+fully-specified utility function may require more
+extensive development and validation or may vary widely across patients.
 

### 2025-06-25 choosing_ordinal_effect_measures.tex
Add Kalbfleisch et al reference

@@
 
+Another set of critiques are in
+\citet{hartmanPitfallsConcordanceIndex2023}. I haven't understood
+these yet but the main concern seems to be that more comparable cases
+(in terms of baseline risk) are more likely to have ties. So there's
+a tradeoff between relevance of the comparisons (conditioning on
+similar baseline risk) and informativeness of the comparisons
+(avoiding ties/getting clear "wins").
+
 Simulations show a typically very strong association between

### 2025-08-10 constructing_ordinal_outcomes.tex
Some restructuring and feedback

@@
 
-\title{Constructing ordinal outcomes for clinical trials}
+\title{Constructing composite ordinal outcomes for clinical trials}
 \author{Leon Di Stefano + Ordinal
@@
 \begin{abstract}
-  Ordinal outcomes are an increasingly popular choice for
-  clinical trials.
-  This paper is a guide to techniques for constructing ordinal
-  outcomes out of pre-existing components, aimed at clinicians and their
-  statistical collaborators.
-  We discuss three broad mechanisms for combining existing binary,
-  ordinal, or continuous
-  outcomes into an ordinal composite---overriding (or
-  ``sequential composition''), tie-breaking (or ``hierarchical
-  composition''), and what we term ``parallel
-  composition''---and illustrate each mechanism with a real-world example.
-  We discuss how these mechanisms can improve the validity and
-  efficiency of ordinal outcomes, but also why coarsening outcomes by
-  combining levels might be preferred in practice.
-  Our approach is closely related to the literature on prioritized
-  outcomes (like win methods and
-    desirability of outcome
-  ranking methodologies), but we focus on the task of
-  constructing ordinal
-  outcomes and distinguish this from the related tasks of choosing
-  effect measures and estimators and
-  handling censoring.
+
+  \todo{Suggested framing of this based on 2025-08-07 meeting with
+    Kate and Rob is: advice on constructing ordinal outcomes from
+    pre-existing components. In particular, ``constructing'' consists
+    of the steps before ``validating'' (using external data). The two
+    main headings are proposed to be: ``combining existing outcomes''
+  and ``how many categories?''}
+
+  Ordinal outcomes are an increasingly popular choice for clinical trials. This
+  paper is a guide to techniques for constructing ordinal outcomes out of
+  pre-existing components, aimed at clinicians and their statistical
+  collaborators. We discuss three broad mechanisms for combining existing
+  ordinal outcomes---including binary, continuous, and time to event
+  outcomes---into an ordinal composite: overriding (or ``sequential
+  composition''), tie-breaking (or ``hierarchical composition''), and what we
+  term ``parallel composition''. We illustrate these mechanisms with examples
+  from the literature and discuss how constructing fine-grained composite
+  outcomes can improve validity and efficiency---but also why coarser outcomes
+  might sometimes be preferred in practice. Finally, we connect these general
+  methods for combining ordinal outcomes with the ``prioritized outcomes''
+  literature (including win-based and desirability of outcome ranking (DOOR)
+  methodologies), clearly delineating the task of constructing composite
+  ordinal outcomes from the related but distinct tasks of choosing effect
+  measures, choosing estimators, and handling censoring.
+
 \end{abstract}
@@
 
-Ordinal outcomes (or ``endpoints'') have become increasingly
-prominent in clinical
-resarch, particularly following the COVID-19 pandemic, when ordinal
-scales developed by the WHO and NIAID were used in prominent trials
-\citep{beigelRemdesivirTreatmentCovid192020,
-marshallMinimalCommonOutcome2020}. Ordinal outcomes have also
-been used in a number of adaptive platform trials
-\citep{angusREMAPCAPRandomizedEmbedded2020,
-walkerCodesigningNovelOrdinal2025}.
-
-Ordinal outcomes rank patient outcomes from better to worse, but
-\emph{differences} in patient outcomes are
-not treated as meaningful.
+Ordinal outcomes (or endpoints) rank patients trajectories from better to
+worse, but do not treat \emph{differences} in individual patient outcomes as
+meaningful. Some ordinal outcomes have a long history in specific disease
+areas, for example the modified Rankin scale in acute stroke trials
+\citep{broderickEvolutionModifiedRankin2017}, and ordinal outcomes have become
+increasingly prominent in clinical resarch following the COVID-19 pandemic when
+disease-specific ordinal outcome scales were developed by the WHO and NIAID
+\citep{beigelRemdesivirTreatmentCovid192020, marshallMinimalCommonOutcome2020}.
+Ordinal outcomes have also been used in a number of adaptive platform trials
+\citep{angusREMAPCAPRandomizedEmbedded2020, walkerCodesigningNovelOrdinal2025}.
 
 Two desirable features of outcomes in general are \emph{validity} and
-\emph{efficiency}.
-\emph{Valid} outcomes reflect
-patient benefit and harm; \emph{efficient} outcomes
-are able to discriminate among better- and worse-off
-patients without large sample sizes.
+\emph{efficiency}. \emph{Valid} outcomes reflect patient benefit and harm;
+\emph{efficient} outcomes are able to discriminate among better- and worse-off
+patients with relatively small sample sizes. (Strictly speaking efficiency is a
+  property not of outcomes themselves but of estimators of effect measures for
+  summarizing differences in outcome distribution---more on this in
+\ref{sec:prioritized_outcomes}.)
 
@@
     safety events into a
-    primary outcome.
+    primary outcome; this might be particularly useful in pragmatic
+    contexts or where side-effects of an intervention are difficult to identify.
 
@@
 
-For the purposes of this
-article, set aside ease of collection, low missingess, objective
-ascertainability, etc.
-
 An example of an outcome which is valid but inefficient might be
@@
 
-The goal of this manuscript is to serve as a guide for clinicians
-and trialists
+The goal of this manuscript is to serve as a guide for clinicians and trialists
 in developing ordinal outcomes for their studies.
 
-Another goal is to situate the literature around ``generalized
-pairwise comparisons of prioritized outcomes''
+Another goal is to situate the literature around ``generalized pairwise
+comparisons of prioritized outcomes''
 \citep{buyseGeneralizedPairwiseComparisons2022}--including win ratio
@@
 ongUnlockingDOORHow2023}--within the broader context of ordinal outcomes and
-endpoints.
+endpoints. These methods as typically presented implicitly bundle
+together 3 distinct things:
+(1) a choice of ordinal outcome or endpoint (composed through
+tie-breaking in the terminology below); (2)
+a choice of effect measure (a.k.a.~target parameter, estimand) based
+on the ``independent pairs'' construction, and (3) a method for
+handling censoring, particularly when the outcomes making up the
+composite include time-to-event information.
 
@@
 
-\paragraph{Outcomes versus effect measures} It is important to
-distinguish ``outcomes'' and ``effect measures'' -- both
-sometimes referred to under the banner of ``endpoints''. Here we
-focus on outcomes; we discuss endpoints in a separate paper \todo{cite}.
+\paragraph{Outcomes versus effect measures} It is important to distinguish
+``outcomes'' and ``effect measures'' -- both sometimes referred to under the
+banner of ``endpoints''. Here we focus on outcomes; we discuss endpoints in a
+separate paper \todo{cite}.
 
 \paragraph{Constructing versus validating} We want to distinguish
-\emph{choosing} or \emph{constructing} an
-outcome/effect
-measure from \emph{validating} the outcome/effect measure using
-external data.
-This paper is about the first task, choosing (or constructing)
-the outcome and
-effect measure. In practice choosing an outcome and effect measure and
-validating that outcome and effect measure in pre-trial data will be an
-iterative process. Our focus in this paper is on the ``positive''
-side of the
-process--namely coming up with, and tweaking, proposed outcomes and effect
-measures--rather than the ``negative'' side, namely testing or checking
-proposals with data to assess whether they have desired properties.
+\emph{choosing} or \emph{constructing} an outcome/effect measure from
+\emph{validating} the outcome/effect measure using external data. This paper is
+about the first task, choosing (or constructing) the outcome and effect
+measure. In practice choosing an outcome and effect measure and validating that
+outcome and effect measure in pre-trial data will be an iterative process. Our
+focus in this paper is on the ``positive'' side of the process--namely coming
+up with, and tweaking, proposed outcomes and effect measures--rather than the
+``negative'' side, namely testing or checking proposals with data to assess
+whether they have desired properties.
 
-\section{Reasons to combine (or fine-grain) ordinal outcomes}
+\subsection{Reasons to use ordinal outcomes}
 
@@
 
-\subsection{Incorporating intercurrent events, especially death,
-for validity}
-
-The intercurrent event could be (1) incorporated as a level (2) used
-for tie-breaking or overriding.
-
-This is a version of the ``composite strategy'' as described in e.g.
-\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
-more efficiency than creating a \emph{binary} composite outcome.
-
-For example overriding a numeric outcome by death.
-Example: REMAP-CAP: organ-free survival, overridden by death
-(Figure \ref{fig:remap_cap_override}).
-
-\begin{figure}
-  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
-  \caption{\textbf{Overriding a continuous outcome, illustrated by
-    REMAP-CAP}. Outcome from the REMAP-CAP adaptive
-    platform trial
-    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
-    Illustrates overriding a numeric ordinal outcome (organ
-    support-free days) by a binary ordinal outcome (death; this is
-    ``sequential composition''), which
-    would also be an intercurrent event. This is another example of
-  sequential composition.}
-  \label{fig:remap_cap_override}
-\end{figure}
-
-\subsection{Increasing efficiency by shrinking the largest category}
-
-Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
-efficiency (variance) of an unadjusted treatment comparison using an ordinal
-versus continuous outcome in the proportional odds model is given
-by $1 - \sum
-\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
-over control and experimental groups) of outcome category $c$.
-Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
-categories when the probabilities are all equal: $\overline p_c = 1/C$.
-
-The results are based on estimating a common odds ratio, but in large
-enough samples are equivalent to the Mann-Whitney-Wilcoxon test, and
-therefore to testing the various ``win nulls'' (win probability is
-  0.5, win difference
-is 0, win ratio is 1).
-
-We have simulation results and some maths to show that the value of $\sum_c
-\overline p_c^3$ -- and therefore the (relative) efficiency of the
-ordinal outcome -- is essentially a function of the largest category
-probability $\overline{p}_{\max} = \max \overline{p}_c$. See Figure
-\ref{fig:p_max} for an illustration of the maths.
-
-\begin{figure}
-  \includegraphics[width=6in]{p_max_controls_efficiency.png}
-  \caption{\textbf{Efficiency gains are controlled by the probability
-    of the largest category.} Bounds on the relative efficiency of an
-    ordinal outcome
-    compared to its continuous counterpart as a function of the maximum
-    category probability $p_\text{max}$. Results are based on
-    estimating a common odds ratio or testing stochastic ordering
-    using the Mann-Whitney-Wilcoxon test. The upper bound (best case
-    relative efficiency for a given $p_\text{max}$) is always
-    achieved in the
-    limit of infinitely many categories ($C = \infty$). The lower bound
-    is always achieved when the number of categories is as small as
-    possible for a given $p_\text{max}$; in particular when
-  $p_\text{max} \geq 0.5$ the worst case occurs with 2 categories.}
-  \label{fig:p_max}
-\end{figure}
-
-Thus simple advice is: for statistical efficiency, try to make
-the proportion
-of patients in the largest category small. However a converse is: if the
-proportion of patients in the largest category is small, not much
-is gained by
-fine-graining; a coarser outcome might be easier to communicate.
-
-\subsection{Augmenting a state or time-to-event outcome with
+\subsubsection{Augmenting a state or time-to-event outcome with
 longitudinal information}
@@
 
-\section{Ways to combine ordinal outcomes}
+\subsection{Incorporating intercurrent events, especially death,
+for validity}
 
-\subsection{Sequential,
-hierarchical, \& parallel composition}
+The intercurrent event could be (1) incorporated as a level (2) used
+for tie-breaking or overriding.
+
+This is a version of the ``composite strategy'' as described in e.g.
+\citet{kahanEstimandsFrameworkPrimer2024}, but we expect it to yield
+more efficiency than creating a \emph{binary} composite outcome.
+
+For example overriding a numeric outcome by death.
+Example: REMAP-CAP: organ-free survival, overridden by death
+(Figure \ref{fig:remap_cap_override}).
+
+\begin{figure}
+  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
+  \caption{\textbf{Overriding a continuous outcome, illustrated by
+    REMAP-CAP}. Outcome from the REMAP-CAP adaptive
+    platform trial
+    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
+    Illustrates overriding a numeric ordinal outcome (organ
+    support-free days) by a binary ordinal outcome (death; this is
+    ``sequential composition''), which
+    would also be an intercurrent event. This is another example of
+  sequential composition.}
+  \label{fig:remap_cap_override}
+\end{figure}
+
+\subsection{Gaining statistical efficiency}
+
+(See below.)
+
+\section{Three ways to create composite ordinal outcomes}
+
+\subsection{Reviewing the literature for existing/core outcomes}
+
+\subsection{Overriding, tie-breaking, \& parallel composition}
 
@@
 
-\subsection{Pairwise comparison methods (win and DOOR) implicitly construct
-ordinal outcomes}
+\section{How many levels?}
+
+\subsection{Increasing efficiency by shrinking the largest category}
+
+Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
+efficiency (variance) of an unadjusted treatment comparison using an ordinal
+versus continuous outcome in the proportional odds model is given
+by $1 - \sum
+\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
+over control and experimental groups) of outcome category $c$.
+Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
+categories when the probabilities are all equal: $\overline p_c = 1/C$.
+
+The results are based on estimating a common odds ratio, but in large
+enough samples are equivalent to the Mann-Whitney-Wilcoxon test, and
+therefore to testing the various ``win nulls'' (win probability is
+  0.5, win difference
+is 0, win ratio is 1).
+
+We have simulation results and some maths to show that the value of $\sum_c
+\overline p_c^3$ -- and therefore the (relative) efficiency of the
+ordinal outcome -- is essentially a function of the largest category
+probability $\overline{p}_{\max} = \max \overline{p}_c$. See Figure
+\ref{fig:p_max} for an illustration of the maths.
+
+\begin{figure}
+  \includegraphics[width=6in]{p_max_controls_efficiency.png}
+  \caption{\textbf{Efficiency gains are controlled by the probability
+    of the largest category.} Bounds on the relative efficiency of an
+    ordinal outcome
+    compared to its continuous counterpart as a function of the maximum
+    category probability $p_\text{max}$. Results are based on
+    estimating a common odds ratio or testing stochastic ordering
+    using the Mann-Whitney-Wilcoxon test. The upper bound (best case
+    relative efficiency for a given $p_\text{max}$) is always
+    achieved in the
+    limit of infinitely many categories ($C = \infty$). The lower bound
+    is always achieved when the number of categories is as small as
+    possible for a given $p_\text{max}$; in particular when
+  $p_\text{max} \geq 0.5$ the worst case occurs with 2 categories.}
+  \label{fig:p_max}
+\end{figure}
+
+Thus simple advice is: for statistical efficiency, try to make
+the proportion
+of patients in the largest category small. However a converse is: if the
+proportion of patients in the largest category is small, not much
+is gained by
+fine-graining; a coarser outcome might be easier to communicate.
+
+\subsection{Consider efficiency within levels of strong predictors}
+
+\subsection{Coarsening to avoid comparing ambiguous levels (validity)}
+
+In the PASSPORT outcome (Figure \ref{fig:passport_outcome}), for
+example, we include both digit
+amputation and ECMO
+any time within 28 days post-enrollment under the heading of
+``adverse clinical events''.
+
+Including these outcomes within the same level of an
+ordinal outcome does not mean assuming that the two outcomes have the
+\emph{same} utility. Rather it represents \emph{avoiding comparing}
+their utilities. (This was something clinicians asked about more than once.)
+
+Also, we "chunk" longer hospital stays more aggressively to avoid
+small categories, for the sake of interpretability.
+
+\subsection{Coarsening for ease of communication}
+
+Too many levels can make it hard to directly compare meaningful
+proportions/probabilities.
+
+(Outocmes with very many levels -- even up to of the order of 100 --
+aren't really a problem for model fitting with modern software.)
+
+Coarsening, including combining levels and replacing complete
+tie-breaks with overrides, could help with this.
+
+\section{Hierarchical ordinal outcomes and pairwise comparison methods}
 
@@
 
-\section{Reasons to coarsen ordinal outcomes}
-
-\subsection{Coarsening to avoid comparing ambiguous levels (validity)}
-
-In the PASSPORT outcome (Figure \ref{fig:passport_outcome}), for
-example, we include both digit
-amputation and ECMO
-any time within 28 days post-enrollment under the heading of
-``adverse clinical events''.
-
-Including these outcomes within the same level of an
-ordinal outcome does not mean assuming that the two outcomes have the
-\emph{same} utility. Rather it represents \emph{avoiding comparing}
-their utilities. (This was something clinicians asked about more than once.)
-
-Also, we "chunk" longer hospital stays more aggressively to avoid
-small categories, for the sake of interpretability.
-
-\subsection{Coarsening for ease of communication}
-
-Too many levels can make it hard to directly compare meaningful
-proportions/probabilities.
-
-(Outocmes with very many levels -- even up to of the order of 100 --
-aren't really a problem for model fitting with modern software.)
-
-Coarsening, including combining levels and replacing complete
-tie-breaks with overrides, could help with this.
+Our approach is closely related to that of
+\citet{follmannAnalysisOrderedComposite2020}, who explicitly consider the
+ordinal outcome implicitly constructed by win methods from a time-to-event
+analysis of ordered events, and show how right-censoring of event times induces
+interval censoring of the cosntructed outcome.
 
@@
 
-Our approach is closely related to that of
-\citet{follmannAnalysisOrderedComposite2020},
-who explicitly consider the ordinal outcome implicitly constructed by
-win methods from a time-to-event analysis of ordered events, and show
-how right-censoring of event times induces interval censoring of the
-cosntructed outcome.
+For the purposes of this article we have set aside other desiderata for
+outcomes like ease of collection, low missingess, objective ascertainability,
+etc.
 
 One point of view is that one would ``ideally'' analyze ordinal outcomes by
-reporting the associated expected utilities \todo{cite Berry
-  talk; example of
-utility-weighted Rankin scale} and that since existing ordinal
-methods supply
-such utilities implicitly why not just supply them explicitly. A
-counterpoint
-is that the methods do not, in fact, supply utilities implicitly.
-Another is that ordinal outcomes let one ``hedge'' about precise
-utillities and this is OK:
-fully-specified utility function may require more
-extensive development and validation or may vary widely across patients.
+reporting the associated expected utilities \todo{cite Berry talk; example of
+utility-weighted Rankin scale} and that since existing ordinal methods supply
+such utilities implicitly why not just supply them explicitly. A counterpoint
+is that the methods do not, in fact, supply utilities implicitly. Another is
+that ordinal outcomes let one ``hedge'' about precise utilities and this is
+OK: fully-specified utility function may require more extensive development and
+validation or may vary widely across patients.
+
+\subsection{Censoring, consistent measurement, etc. are all
+additional aspects of constructing the outcome that need to be considered}
 

### 2025-09-15 constructing_ordinal_outcomes.tex
Tweaking

@@
 
+This should also hold \emph{within strata}.
+
 \subsection{Consider efficiency within levels of strong predictors}
@@
 One point of view is that one would ``ideally'' analyze ordinal outcomes by
-reporting the associated expected utilities \todo{cite Berry talk; example of
-utility-weighted Rankin scale} and that since existing ordinal methods supply
-such utilities implicitly why not just supply them explicitly. A counterpoint
-is that the methods do not, in fact, supply utilities implicitly. Another is
-that ordinal outcomes let one ``hedge'' about precise utilities and this is
-OK: fully-specified utility function may require more extensive development and
-validation or may vary widely across patients.
+reporting the associated expected utilities \todo{cite Berry talk; possibly
+cite Lumley blog post; example of utility-weighted Rankin scale} and that since
+existing ordinal methods supply such utilities implicitly why not just supply
+them explicitly. A counterpoint is that the methods do not, in fact, supply
+utilities implicitly. Another is that ordinal outcomes let one ``hedge'' about
+precise utilities and this is OK: fully-specified utility function may require
+more extensive development and validation or may vary widely across patients.
 

### 2025-11-06 1_constructing_ordinal_outcomes.tex
A bunch of work

--- 2025-09-15 constructing_ordinal_outcomes.tex
@@
 
-\title{Constructing composite ordinal outcomes for clinical trials}
+\title{Advice on constructing composite ordinal outcomes}
 \author{Leon Di Stefano + Ordinal
@@
   ordinal outcomes---including binary, continuous, and time to event
-  outcomes---into an ordinal composite: overriding (or ``sequential
-  composition''), tie-breaking (or ``hierarchical composition''), and what we
-  term ``parallel composition''. We illustrate these mechanisms with examples
-  from the literature and discuss how constructing fine-grained composite
-  outcomes can improve validity and efficiency---but also why coarser outcomes
-  might sometimes be preferred in practice. Finally, we connect these general
-  methods for combining ordinal outcomes with the ``prioritized outcomes''
-  literature (including win-based and desirability of outcome ranking (DOOR)
-  methodologies), clearly delineating the task of constructing composite
-  ordinal outcomes from the related but distinct tasks of choosing effect
-  measures, choosing estimators, and handling censoring.
+  outcomes---into an ordinal composite: (1) overriding (or ``sequential
+  composition''), (2) tie-breaking (or ``hierarchical composition''), and what
+  we term (3) ``parallel composition''. We illustrate each of these mechanisms
+  with examples from the literature. Next, we discuss how constructing
+  fine-grained composite outcomes can improve validity and efficiency---but
+  also why adding levels sometimes has very little effect on efficiency.
+  Finally, we connect these general methods for combining ordinal outcomes with
+  the ``prioritized outcomes'' literature---including win-based and
+  desirability of outcome ranking (DOOR) methodologies---clearly delineating
+  the task of constructing composite ordinal outcomes from the related but
+  distinct tasks of choosing effect measures, choosing estimators, and handling
+  censoring.
 
@@
 worse, but do not treat \emph{differences} in individual patient outcomes as
-meaningful. Some ordinal outcomes have a long history in specific disease
-areas, for example the modified Rankin scale in acute stroke trials
-\citep{broderickEvolutionModifiedRankin2017}, and ordinal outcomes have become
-increasingly prominent in clinical resarch following the COVID-19 pandemic when
-disease-specific ordinal outcome scales were developed by the WHO and NIAID
-\citep{beigelRemdesivirTreatmentCovid192020, marshallMinimalCommonOutcome2020}.
+meaningful. Ordinal outcomes have a long history in specific disease
+areas---for example the modified Rankin scale in stroke trials---and have
+become increasingly prominent in clinical resarch following the COVID-19
+pandemic when disease-specific ordinal outcome scales were developed by the WHO
+and NIAID \citep{broderickEvolutionModifiedRankin2017,
+beigelRemdesivirTreatmentCovid192020, marshallMinimalCommonOutcome2020}.
 Ordinal outcomes have also been used in a number of adaptive platform trials
@@
 Two desirable features of outcomes in general are \emph{validity} and
-\emph{efficiency}. \emph{Valid} outcomes reflect patient benefit and harm;
-\emph{efficient} outcomes are able to discriminate among better- and worse-off
-patients with relatively small sample sizes. (Strictly speaking efficiency is a
-  property not of outcomes themselves but of estimators of effect measures for
-  summarizing differences in outcome distribution---more on this in
-\ref{sec:prioritized_outcomes}.)
+\emph{efficiency}. \emph{Valid} outcomes reflect differences in patient benefit
+and harm; \emph{efficient} outcomes are able to discriminate among better- and
+worse-off patients with relatively small sample sizes. (Each of these is a
+function of other factors too.)
+
+Ordinal outcomes have the potential to
 
@@
 \end{itemize}
-
-An example of an outcome which is valid but inefficient might be
-death as a binary outcome in a context where deaths
-are rare.
-
-An example of an outcome which is efficient but which may be invalid
-might be continuously-measured concentrations of a biomarker which is a poor
-surrogate for disease progression, or a continuous outcome that
-excludes mortality.
 
@@
 
-\section{Three ways to create composite ordinal outcomes}
+\section{Three ways to combine existing ordinal scales}
 
@@
 
-\subsection{Increasing efficiency by shrinking the largest category}
+\subsection{More levels: increasing efficiency by
+shrinking the largest category proportion}
 
@@
 
-\subsection{Consider efficiency within levels of strong predictors}
+\subsection{More levels: Consider efficiency within levels of strong predictors}
 
-\subsection{Coarsening to avoid comparing ambiguous levels (validity)}
+\begin{figure}
+  \includegraphics[width=7in]{stratified_efficiency_plots.png}
+  \caption{\textbf{Overriding (sequential composition) illustrated
+    with the PASSPORT outcome} Comparison of the ordinal outcome used
+  in PASSPORT (depicted in Figure \ref{fig:passport_outcome})} with a
+  coarsened 4-level version, stratified by Australisian Triage Scale category.
+  \label{fig:stratified_efficiency}
+\end{figure}
+
+\subsection{Fewer levels: avoid comparing ambiguous levels (validity)}
 
@@
 
-\subsection{Coarsening for ease of communication}
+\subsection{Fewer levels: ease of communication}
 
@@
 
-\section{Hierarchical ordinal outcomes and pairwise comparison methods}
+\section{Composite ordinal outcomes and ``win'' methods}
 
